---
title: "Topic 1-03: MSigDB gene sets"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-03: MSigDB gene sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6,
    fig.height = 6,
    fig.align = "center"
)
rm(list = ls())
```

## The .gmt data format

MSigDB defines a simple `.gmt` format for storing gene sets. It has the following format:

```
gene_set_1  gene_set_description    gene1   gene2   gene3
gene_set_2  gene_set_description    gene4   gene5
...
```

The `.gmt` format is used more and more for sharing gene set data, e.g. https://maayanlab.cloud/Enrichr/#libraries.


Let's try to read a `.gmt` file. We use the `.gmt` file for the Hallmark gene set.

```{r}
download.file("https://data.broadinstitute.org/gsea-msigdb/msigdb/release/2024.1.Hs/h.all.v2024.1.Hs.symbols.gmt",
    destfile = "h.all.v2024.1.Hs.symbols.gmt")
```

Since it is not a table, we have to read the file by lines.

```{r}
ln = readLines("h.all.v2024.1.Hs.symbols.gmt")
ln = strsplit(ln, "\t")
gs = lapply(ln, function(x) x[-(1:2)])
gs_names = sapply(ln, function(x) x[1])
names(gs) = gs_names
gs[1:2]

gs_description = sapply(ln, function(x) x[2])
names(gs_description) = gs_names
gs_description[1:2]
```


```{r, echo = FALSE}
invisible(file.remove("h.all.v2024.1.Hs.symbols.gmt"))
```

## Read gene sets from MSigDB

All the MSigDB gene sets are hosted on https://data.broadinstitute.org/gsea-msigdb/msigdb/release/.
In the **GSEAtopics** packages, there are several helper functions to get gene sets from a specific collection.

```{r}
library(GSEAtopics)
list_msigdb_versions()
list_msigdb_collections("2024.1.Hs")
lt = get_msigdb(version = "2024.1.Hs", collection = "h.all")
lt[1:2]
```

## Other package

There is a package **msigdbr** on https://cran.r-project.org/web/packages/msigdbr/index.html. In the **rGREAT** package,
there is al `read_gmt()` function which reads from a `.gmt` file and supports gene ID conversion.

## Practice 

### Practice 1 {.tabset}

#### Practice

The gene set resource on https://maayanlab.cloud/Enrichr/#libraries is very useful. You may
want to use it some day in the future. Take one gene set collection (e.g. the COVID-19 related gene sets), download the corresponding gene set file (in `.gmt` format) and try to read into R as a list or a two-column data frame.

#### Solution


```{r}
download.file("https://maayanlab.cloud/Enrichr/geneSetLibrary?mode=text&libraryName=COVID-19_Related_Gene_Sets", destfile = "covid-19.gmt")
ln = readLines("covid-19.gmt")
ln = strsplit(ln, "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])
gs[1:2]
```

Convert `gs` to a data frame

```{r}
df = data.frame(
    geneset = rep(names(gs), times = sapply(gs, length)),
    gene = unlist(gs)
)
head(df)
```

Or use the `list_to_data_frame()` function in **GSAEtopics**:

```{r}
df = list_to_data_frame(gs)
head(df)
```

```{r, echo = FALSE}
invisible(file.remove("covid-19.gmt"))
```

