---
title: "Topic 5-02: Cluster and summarize"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 5-02: simplifyEnrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```

There are two types of similarity methods, the count-based and semantic-based. We look at them separately.


## Count-based similarities

As an example, we use significant gene sets from ORA on the p53 dataset.

```{r}
lt = readRDS(system.file("extdata", "p53_expr.rds", package = "GSEAtutorial"))
expr = lt$expr
condition = lt$condition

ln = strsplit(readLines(system.file("extdata", "c2.symbols.gmt", package = "GSEAtutorial")), "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])

library(genefilter)
tdf = rowttests(expr, factor(condition))
tdf$fdr = p.adjust(tdf$p.value, "BH")

diff = rownames(expr)[abs(tdf$statistic) > 2]
tb_ora = ora(diff, universe = rownames(expr), gene_sets = gs)
sig = tb_ora$gene_set[tb_ora$p_value < 0.05]
```

We suggest to only use the significant genes when calculating the similarities.

```{r}
gs2 = lapply(gs[sig], function(x) intersect(x, diff))
```

Additionally adding the gene weight can sometimes improve the similarity patterns.

```{r}
w = tdf$statistic
names(w) = rownames(tdf)

sm = jaccard(gs2, weight = abs(w))
diag(sm) = NA
library(ComplexHeatmap)
Heatmap(sm)
```


The block-pattern is very obvious. You can split it by hierarchical clustering + cutree, kmeans or any clustering methods.
We recommend to use the graph-based clustering methods.

The similarity matrix can be thought of as an adjacency matrix. The **igraph** package provides many graph-based clustering methods.

```{r}
library(igraph)
g = graph_from_adjacency_matrix(sm, weighted = TRUE, mode = "upper", diag = FALSE)
cl = membership(cluster_louvain(g))
Heatmap(sm, row_split = cl, column_split = cl)
```

Clustering the gene sets is fairly easy. However, this is not the end of the analysis. We need
to summarize the commonness of genesets in a cluster. Note the similarity is count-based, then
these genesets must share some common genes, i.e., the driving genes that cluster the genesets.

Taking the geneset cluster 1 as an example, which is related to "P53"-related functions.

```{r}
gs2[cl == 1]
```

Let's calculate the recurrency of genes in these gene sets:

```{r}
tb = table(unlist(gs2[cl == 1]))
sort(tb, decreasing = TRUE)
```

We can see "CDK1A" is in all 8 genesets, "MDM2" is in 7 out of 8 gene sets, and "BAX" is in 6 out of 8 gene sets. So
these three genes are driving genes of the eight genesets in this cluster.


For the visualization purpose, can we add such driving genes to the heatmap?

Let's first adjust the heatmap:

```{r}
ht = Heatmap(sm, row_split = cl, column_split = cl, row_names_side = "left",
    show_row_dend = FALSE, show_column_dend = FALSE, row_title = NULL, column_title = NULL)
```

Driving genes are added as a "link" annotation:

```{r, fig.width = 10}
ht + rowAnnotation(gene = anno_link(cl, panel_fun = function(index) {
    tb = table(unlist(gs2[index]))
    tb = sort(tb, decreasing = TRUE)
    pushViewport(viewport())
    grid.rect()
    grid.text(paste(names(tb)[tb/length(index) > 0.5], collapse = "\n"))
    popViewport()
}))
```

In the **GSEAtutorial** package, there is a `genesets_similarity_heatmap()` function.

```{r, fig.width = 10}
genesets_similarity_heatmap(gs2, weight = w)
```

**Question**: in the previous example, the enrichment is done by ORA which is based on gene counts. The similarity is also based on gene counts.
What if the enrichment is done by GSEA? How to adjust the calculation process?

We first apply the GSEA analysis.

```{r}
library(matrixStats)
m1 = expr[, condition == "WT"]  # only samples in group 1
m2 = expr[, condition == "MUT"]  # only samples in group 2
s = (rowMeans(m2) - rowMeans(m1))/(rowSds(m1) + rowSds(m2)) # gene-level difference scores
s = sort(s, decreasing = TRUE)
tb_gsea = fgsea_wrapper(s, gs)
head(tb_gsea)
```

The similarity in this category is based on gene counts. Although the enrichment procedure is not gene count-based. In the results,
GSEA reports "leading edge genes" which are the genes that contribute most to the enrichment of the gene set. These leading-edge genes
can be used for calculating the similarities.

```{r, fig.width = 10}
l = tb_gsea$p_adjust < 0.1
gs3 = tb_gsea[l, "leading_edge"]
names(gs3) = tb_gsea$gene_set[l]
genesets_similarity_heatmap(gs3, weight = s, resolution = 2)
```

**Thoughts**: GSVA or the single-sample GSEA converts the original gene-sample matrix to geneset-sample matrix. One the aim of such transformation
is to find "co-expressed gene sets". As such "co-expression" is actually due to common genes in the gene sets, we can also do such driving-gene analysis
and associated to a group of "co-expressed" genes.

## Semantic similarity

**simplifyEnrichment** is suggested to work on GO enrichment results.

Let's generate 500 random GO IDs:

```{r, echo = FALSE}
library(simplifyEnrichment)
se_opt$verbose = FALSE
```

```{r, fig.width = 8, fig.height = 4.5}
library(simplifyEnrichment)
set.seed(888)
go_id = random_GO(500)
sm = GO_similarity(go_id)
```

The default semantic similarity method is "Sim_XGraSM_2013", which is an adjusted Lin-1998 method, where
$IC(MICA)$ replaced with $\mathrm{avg} IC(CA)$, i.e. average non-zero IC of all common ancestors.


```{r}
Heatmap(sm, col = c("white", "red"))
```

The heatmap shows very clear block patterns on the diagonal. We next try several clustering methods to
see whether they can capture such patterns.

We implement a helper function `test_clustering()` which accepts a cluster label vector and draws the heatmap.

```{r}
test_clustering = function(cl) {
    col_fun = colorRamp2(c(0, quantile(sm, 0.99)), c("white", "red"))
    Heatmap(sm, col = col_fun, border = "black", show_row_names = FALSE, show_column_names = FALSE,
        show_row_dend = FALSE, show_column_dend = FALSE, row_title = NULL, column_title = NULL,
        row_split = cl, column_split = cl)
}
```

We try hierarchical cluster, k-means, graph-based method and binary cut.

```{r}
cl = cutree(hclust(dist(sm)), k = 4)
test_clustering(cl)

cl = kmeans(dist(sm), centers = 4)$cluster
test_clustering(cl)

g = graph_from_adjacency_matrix(sm, weighted = TRUE, mode = "upper", diag = FALSE)
cl = membership(cluster_louvain(g))
test_clustering(cl)

cl = binary_cut(sm)
test_clustering(cl)

cl = as.character(cl)
tb = table(cl)
cl[cl %in% names(tb[tb < 5])] = "others"
test_clustering(cl)
```

As GO terms in a cluster show high semantic similarities, where they should be concentrated in a certain GO branch. The general
biological functions for a GO cluster is summarized by keywords and visualized by word cloud.

```{r, fig.height = 5}
simplifyGO(go_id)
```

## Multiple enrichment results

We follow the standard workflow for unsupervised cluster analysis. We use the Golub ALL dataset and the **cola** package for analysis.

```{r, message = FALSE, fig.width = 8, fig.height = 7}
library(cola)

library(golubEsets)
data(Golub_Merge)
m = exprs(Golub_Merge)

anno = pData(Golub_Merge)

m[m <= 1] = NA
m = log10(m)

m = adjust_matrix(m)

library(preprocessCore)
cn = colnames(m)
rn = rownames(m)
m = normalize.quantiles(m)
colnames(m) = cn
rownames(m) = rn

res = consensus_partition(m, top_value_method = "ATC", partition_method = "skmeans", verbose = FALSE)
```

We know the best number of subgroups is three. We look for signatures in the three-subgroup classification.

```{r}
df = get_signatures(res, k = 3)
```

Now we have four groups of signatures genes. The next analysis is to compare enriched GO functions in the four
lists of signature genes. The corresponding biological question is do these different signature genes have
similar function or they exhibit specific functions?

We first obtain the four list of signature genes.

```{r}
gl = split(df$which_row, df$km)
gl = lapply(gl, function(ind) rownames(res)[ind])
```

The ALL data is almost 25 years ago. The gene ID is Affymetric microarray probe ID, Luckily, the corresponding
annotation package is also available on Bioconductor which is built upon the standard **AnnotationDbi** package.

```{r}
library(hu6800.db)
gl = lapply(gl, function(x) {
    x2 = mapIds(hu6800.db, keys = x, keytype = "PROBEID", column = "ENTREZID")
    unique(x2[!is.na(x2)])
})
```

We aply ORA GO analysis.

```{r}
tbl = lapply(gl, ora_go)
```

`tbl` is a list of four enrichment result tables. 
`simplifyGOFromMultipleLists()` can be used to visualize GO enrichment from multiple results.
The columns for GO IDs and p-values are automatically detected.

```{r, fig.width = 12, fig.height = 6}
simplifyGOFromMultipleLists(tbl, padj_cutoff = 0.001)
```

## Word cloud

In some cases, the similarity heatmap is not necessary to put into the final report.

```{r, fig.width = 8, fig.height = 4.5}
df = tbl[[1]]
go_id = df$gene_set[df$p_adjust < 0.01]
simplifyGO(go_id)
```


`summarizeGO()` makes an even simpler plot with word cloud and simple statistical graphics.

When there is only one enrichment results, there are two inputs for `summarizeGO()`:

- A vector of GO IDs
- A numeric vector of corresponding statistics which are mapped to barplots.

```{r, fig.width = 8, fig.height = 5}
l = df$p_adjust < 0.01
summarizeGO(df$gene_set[l], -log10(df$p_adjust)[l], axis_label = "average -log10(p.adjust)")
```


Or visualize average log2 fold enrichment:

```{r, fig.width = 8, fig.height = 5}
l = df$p_adjust < 0.01
summarizeGO(df$gene_set[l], df$log2fe[l], axis_label = "average log2(fold enrichment)")
```


`summarizeGO()` also supports multiple enrichment results. In this case, `value` should be 
 is a list of numeric named vectors which contains significant GO terms in each enrichment table.

```{r, fig.width = 8, fig.height = 7}
value = lapply(tbl, function(df) {
    v = -log10(df$p_adjust)
    names(v) = df$gene_set
    v[df$p_adjust < 0.001]
})
value
summarizeGO(value = value, axis_label = "average -log10(p_adjust)", 
    legend_title = "-log10(p_adjust)")
```


Or use log2 fold enrichment:

```{r, fig.width = 8, fig.height = 7}
value = lapply(tbl, function(df) {
    v = df$log2fe
    names(v) = df$gene_set
    v[df$p_adjust < 0.001]
})
summarizeGO(value = value, axis_label = "average log2_fold_enrichment", 
    legend_title = "log2(fold_enrichment)")
```



## Practice 

### {.tabset}

#### Practice 1

Recall in [Topic 2-02: Implement ORA from stratch](topic2_02_implement_ora.html), we compared
the ORA results when setting DE cutoffs to 0.05, 0.01 and 0.001. Now perform a simplify enrichment
analysis on the three ORA tables and compare them.

```{r}
de = readRDS(system.file("extdata", "de.rds", package = "GSEAtraining"))
de = de[, c("symbol", "p_value")]
de = de[!is.na(de$p_value), ]

de_genes_1 = de$symbol[de$p_value < 0.05]
de_genes_2 = de$symbol[de$p_value < 0.01]
de_genes_3 = de$symbol[de$p_value < 0.001]

library(GSEAtraining)
library(org.Hs.eg.db)
gs = get_GO_gene_sets_from_orgdb(org.Hs.eg.db, "BP", gene_id_type = "SYMBOL")

tb1 = ora(de_genes_1, gs)
tb2 = ora(de_genes_2, gs)
tb3 = ora(de_genes_3, gs)
```


#### Solution

```{r, fig.width = 12, fig.height = 7}
lt = list(
    "DE_0.05" = tb1,
    "DE_0.01" = tb2,
    "DE_0.001" = tb3
)

simplifyGOFromMultipleLists(lt, padj_cutoff = 0.05)
```

