---
title: "Topic 2-02: Implement ORA from stratch"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-02: Implement ORA from stratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```

Implementing ORA is very simple. As long as genes and gene sets are already in
the correct ID types, the implementation of ORA only needs
a few lines of code.

## Calculate the statistics

There are three input parameters for ORA:

- a vector of (DE) genes: `genes`
- a gene set: `gene_set`
- a vector of background genes: `universe`

With these three objects, we calculate the following statistics:

The number of total genes: 

```r
n_universe = length(universe)
```

The number of DE genes:

```r
k = length(genes)
```

The number of genes in each gene set, i.e., the size of each gene set:

```r
m = length(gene_set)
```

And, the number of DE genes in gene sets:

```r
x = length(intersect(gene_set, genes))
```



<style>
.table {
width:400px;
}
</style>

We put these variables into the 2x2 contigency table:

|      | in the set | not in the set| total |
|------|------------:|---------------:|-------:|
|DE    |  `x`    |   -   |  `k` |
|not DE|  -    |      -   |  - |
|total |  `m`    |    -     |  `n_universe` |


Then we can calculate the p-value from the hypergeometric distribution as:

```r
phyper(x - 1, m, n_universe - m, k, lower.tail = FALSE)
```

which is the same as:

```r
phyper(x - 1, k, n_universe - k, m, lower.tail = FALSE)
```

Note, `phyper()` calculates $P(X > x)$, thus we write `phyper(x - 1, ...)` which corresponds to $P(X > x - 1) = P(X \ge x)$.

## Wrap into a function

A quick and dirty way by first implementing a function working on single gene set, then applying it to a list of gene sets.

```{r}
ora_single = function(genes, gene_set, universe) {
    n_universe = length(universe)
    k = length(genes)
    
    x = length(intersect(gene_set, genes))
    m = length(gene_set)
    n = n_universe - m
    
    p = phyper(x - 1, m, n, k, lower.tail = FALSE)

    p
}
ora =  function(genes, gene_sets, universe) {
    ns = length(gene_sets)
    p = numeric(ns)
    for(i in seq_along(gene_sets)) {
        p[i] = ora_single(genes, gene_sets[[i]], universe)
    }
    p
}
```

Or we can do it in a vectorized way:

```{r}
ora = function(genes, gene_sets, universe) {
    n_universe = length(universe)
    k = length(genes)
    
    x = sapply(gene_sets, function(x) length(intersect(x, genes)))
    m = sapply(gene_sets, length)
    n = n_universe - m
    
    p = phyper(x - 1, m, n, k, lower.tail = FALSE)

    data.frame(
        gene_set = names(gene_sets),
        n_hits = x,
        n_genes = k,
        n_gs = m,
        n_total = n_universe,
        p_value = p
    )
}
```


We can further improve the function by:

1. setting the default of `universe` to the total genes in `gene_sets`,
2. intersecting `genes` and `gene_sets` to `universe`. Note this step also removes duplicated genes,
3. adding log2 fold enrichment and adjusted p-values to the result data frame,
4. ordering the result data frame by the significance.

```{r}
ora = function(genes, gene_sets, universe = NULL) {

    if(is.null(universe)) {
        universe = unique(unlist(gene_sets))
    } else {
        universe = unique(universe)
    }

    genes = intersect(genes, universe)
    gene_sets = lapply(gene_sets, function(x) intersect(x, universe))

    n_universe = length(universe)
    k = length(genes)
    
    x = sapply(gene_sets, function(x) length(intersect(x, genes)))
    m = sapply(gene_sets, length)
    n = n_universe - m
    
    p = phyper(x - 1, m, n, k, lower.tail = FALSE)

    df = data.frame(
        gene_set = names(gene_sets),
        n_hits = x,
        n_genes = k,
        n_gs = m,
        n_universe = n_universe,
        log2fe = log2(x*n_universe/k/m),
        p_value = p
    )

    df$p_adjust = p.adjust(df$p_value, "BH")
    df[order(df$p_adjust, df$p_value), ,drop = FALSE]
}
```

Let's try our `ora()` with 1000 random genes on the GO gene sets:

```{r}
library(org.Hs.eg.db)
library(GSEAtutorial)
gs = get_GO_gene_sets_from_orgdb(org.Hs.eg.db, gene_id_type = "ENTREZID")
gs = lapply(gs, unique)  # important!, there are duplicated genes in some GO terms
genes = random_genes(org.Hs.eg.db, 1000, "ENTREZID")

df = ora(genes, gs)
head(df)
```


## Further optimization

We first rename `ora()` to `ora_old()` later for comparison.

```{r}
ora_old = ora
```

`ora()` is already very fast. However, in the previous implementation, there is still a part which is relatively slow in the function.
Let't first do a profiling.

```{r}
Rprof()
df = ora(genes, gs)
Rprof(NULL)
summaryRprof()
```

We can see the `intersect()` function consumes more than 90% of the total runtime. `intersect()` internally uses the `match()` function,
and there is a faster implementation in the **fastmatch** package. We also optimize the use of the `unlist()` function by adding
one additional argument `use.names = FALSE` since we do not care about the names of the unlisted vector.

```{r}
library(fastmatch)
ora = function(genes, gene_sets, universe = NULL) {

    if(is.null(universe)) {
        universe = unique(unlist(gene_sets, use.names = FALSE))
    } else {
        universe = unique(universe)
    }
    
    genes = intersect(genes, universe)
    gene_sets = lapply(gene_sets, function(x) x[x %fin% universe])

    n_universe = length(universe)
    k = length(genes)
    
    x = sapply(gene_sets, function(x) sum(x %fin% genes))
    m = sapply(gene_sets, length)
    n = n_universe - m
    
    p = phyper(x - 1, m, n, k, lower.tail = FALSE)

    df = data.frame(
        gene_set = names(gene_sets),
        n_hits = x,
        n_genes = k,
        n_gs = m,
        n_total = n_universe,
        log2fe = log2(x*n_universe/k/m),
        p_value = p
    )

    df$p_adjust = p.adjust(df$p_value, "BH")
    df[order(df$p_adjust, df$p_value), ,drop = FALSE]
}
```

Now let's benchmark `ora_old()` and the new `ora()`:

```{r}
library(microbenchmark)
microbenchmark(
    ora_old = ora_old(genes, gs),
    ora = ora(genes, gs),
    times = 10
)
```

`ora()` is more than 40x faster than `ora_old()`! We have a very fast implementation of ORA 
than majority of others from R/Bioconductor.

## More

There is still 5% ~ 10% space for further optimization. Note in the previous `ora()`, 
`x %fin% universe` actually matches between two character vectors (internally by hash tables).
If we can change it to match between two integer vectors, it might further improve the code.

In the next improved implementation, we still use `fmatch()` to perform fast matching. We change
genes in objects `genes` and `gene_sets` all to the corresponding integer indicies in `universe`.
This is done by `fmatch()`.

```{r}
ora2 = function(genes, gene_sets, universe = NULL) {

    if(is.null(universe)) {
        universe = unique(unlist(gene_sets, use.names = FALSE))
    } else {
        universe = unique(universe)
    }

    # convert genes to integer indicies
    genes = fmatch(genes, universe); genes = genes[!is.na(genes)]
    gene_sets = lapply(gene_sets, function(x) {
        v = fmatch(x, universe)
        v[!is.na(v)]
    })

    n_universe = length(universe)
    k = length(genes)
    
    x = sapply(gene_sets, function(x) sum(x %fin% genes)) # now %fin% is applied on integer vectors
    m = sapply(gene_sets, length)
    n = n_universe - m
    
    p = phyper(x - 1, m, n, k, lower.tail = FALSE)

    df = data.frame(
        gene_set = names(gene_sets),
        n_hits = x,
        n_genes = k,
        n_gs = m,
        n_total = n_universe,
        log2fe = log2(x*n_universe/k/m),
        p_value = p
    )

    df$p_adjust = p.adjust(df$p_value, "BH")
    rownames(df) = df$gene_set
    df[order(df$p_adjust, df$p_value), ,drop = FALSE]
}
```

```{r}
microbenchmark(
    ora  = ora(genes, gs),
    ora2 = ora2(genes, gs),
    times = 10
)
```

Let's make a final version of `ora()`:


```{r}
ora = function(genes, gene_sets, universe = NULL, min_hits = 3, min_size = 5, max_size = 2500) {

    if(is.null(universe)) {
        universe = unique(unlist(gene_sets, use.names = FALSE))
    } else {
        universe = unique(universe)
    }

    # convert genes to integer indicies
    genes = fmatch(genes, universe); genes = genes[!is.na(genes)]
    gene_sets = lapply(gene_sets, function(x) {
        v = fmatch(x, universe)
        v[!is.na(v)]
    })

    nl = sapply(gene_sets, length)
    gene_sets = gene_sets[nl >= min_size & nl <= max_size]

    n_universe = length(universe)
    k = length(genes)
    
    x = sapply(gene_sets, function(x) sum(x %fin% genes)) # now %fin% is applied on integer vectors
    m = sapply(gene_sets, length)
    n = n_universe - m
    
    p = phyper(x - 1, m, n, k, lower.tail = FALSE)

    df = data.frame(
        gene_set = names(gene_sets),
        n_hits = x,
        n_genes = k,
        n_gs = m,
        n_total = n_universe,
        log2fe = log2(x*n_universe/k/m),
        p_value = p
    )

    df = df[df$n_hits >= min_hits, , drop = FALSE]

    df$p_adjust = p.adjust(df$p_value, "BH")
    df = df[order(df$p_adjust, df$p_value), ,drop = FALSE]
    rownames(df) = NULL
    df
}
```

## Practice 

### {.tabset}

#### Practice 1

The file `de.rds` contains results from a differential expression (DE) analysis.

```{r}
de = readRDS(system.file("extdata", "de.rds", package = "GSEAtutorial"))
de = de[, c("symbol", "p_value")]
de = de[!is.na(de$p_value) & de$symbol != "", ]
head(de)
```

Take the `symbol` and `p_value` columns, set cutoff of `p_value` to 0.05, 0.01
and 0.001 respectively to filter the significant DE genes (of course, in
applications, you need to correct p-values), apply `ora()` to the three DE
gene lists using the GO BP gene sets (think how to obtain the GO gene sets from **org.Hs.eg.db**) and compare the ORA results (the whatever comparison
you can think of).

If you EntreZ ID in the gene sets, you need to convert gene symbols for the DE genes to EntreZ IDs.
But you can also choose to use gene symbols in the gene sets.

#### Solution

The three DE gene lists:

```{r}
de_genes_1 = de$symbol[de$p_value < 0.05]
de_genes_2 = de$symbol[de$p_value < 0.01]
de_genes_3 = de$symbol[de$p_value < 0.001]

length(de_genes_1)
length(de_genes_2)
length(de_genes_3)
```

Get the GO BP gene sets. Note genes in `de_genes_*` are symbols, thus genes in gene sets
should also be symbols.

```{r}
gs = get_GO_gene_sets_from_orgdb(org.Hs.eg.db, "BP", gene_id_type = "SYMBOL")
gs[1]
```


We perform ORA for the three DE gene lists.

```{r}
tb1 = ora(de_genes_1, gs)
tb2 = ora(de_genes_2, gs)
tb3 = ora(de_genes_3, gs)
```

Compare number of significant GO terms:

```{r}
sum(tb1$p_adjust < 0.05)
sum(tb2$p_adjust < 0.05)
sum(tb3$p_adjust < 0.05)
```

Actually we can see gene set enrichment analysis is a robust method
which is not sensitive to DE cutoffs (at least for this dataset).

Make a Venn (Euler) diagram:


```{r}
library(eulerr)
plot(euler(list("DE_cutoff_0.05"  = tb1$gene_set[tb1$p_adjust < 0.05],
                "DE_cutoff_0.01"  = tb2$gene_set[tb2$p_adjust < 0.05],
                "DE_cutoff_0.001" = tb3$gene_set[tb3$p_adjust < 0.05])), 
     quantities = TRUE)
```

And the Venn diagram for DE genes:

```{r}
plot(euler(list("DE_cutoff_0.05"  = de_genes_1,
                "DE_cutoff_0.01"  = de_genes_2,
                "DE_cutoff_0.001" = de_genes_3)), 
     quantities = TRUE)
```



### {.tabset}

#### Practice 2

The `ora()` function can be used as the core part of a full-functional function for ORA analysis.
Please implement a `ora_go()` for GO analysis


#### Solution


```{r}
CACHE = new.env(parent = emptyenv())

load_go_genesets = function(org_db, ontology) {
    db_info = dbInfo(dbconn(org_db))
    organism = db_info$value[db_info$name == "ORGANISM"]
    if(is.null(CACHE$go_genesets[[organism]][[ontology]])) {
        suppressMessages(tb <- AnnotationDbi::select(org_db, keys = ontology, keytype = "ONTOLOGYALL", columns = c("ENTREZID", "GOALL")))
        gs = split(tb$ENTREZID, tb$GOALL)
        go_gs = lapply(gs, function(x) {
            as.character(unique(x))
        })

        go_names = Term(GOTERM)[names(go_gs)]

        CACHE$go_genesets[[organism]][[ontology]]$go_names = go_names
        CACHE$go_genesets[[organism]][[ontology]]$go_gs = go_gs
    } else {
        go_names = CACHE$go_genesets[[organism]][[ontology]]$go_names
        go_gs = CACHE$go_genesets[[organism]][[ontology]]$go_gs
    }

    list(names = go_names, gs = go_gs)
}

ora_go = function(genes, org_db = org.Hs.eg.db::org.Hs.eg.db, ontology = "BP", ...) {
    lt = load_go_genesets(org_db, ontology)
   
    df = ora(genes, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}

load_kegg_genesets = function(organism, db) {
    if(is.null(CACHE$kegg_pathways[[db]][[organism]])) {
        if(db == "pathway") {
            gs_names = read.table(url(paste0("https://rest.kegg.jp/list/", db, "/", organism)), sep = "\t", quote = "")
        } else {
            gs_names = read.table(url(paste0("https://rest.kegg.jp/list/", db)), sep = "\t", quote = "")
        }
        gs_names = structure(gs_names[, 2], names = gs_names[, 1])

        gs_genes = read.table(url(paste0("https://rest.kegg.jp/link/", db, "/", organism)), sep = "\t", quote = "")
        gs_genes[, 1] = gsub("^\\w+:", "", gs_genes[, 1])
        gs_genes[, 2] = gsub("^\\w+:", "", gs_genes[, 2])
        if(db == "module") {
            gs_genes[, 2] = gsub("^[a-zA-Z]+_", "", gs_genes[, 2])
        }

        gs = split(gs_genes[, 1], gs_genes[, 2])
        gs = lapply(gs, function(x) {
            as.character(unique(x))
        })

        CACHE$kegg_pathways[[db]][[organism]]$gs_names = gs_names
        CACHE$kegg_pathways[[db]][[organism]]$gs = gs
    } else {
        gs_names = CACHE$kegg_pathways[[db]][[organism]]$gs_names
        gs = CACHE$kegg_pathways[[db]][[organism]]$gs
    }

    list(names = gs_names, gs = gs)
}

# db: pathway, module, ko, network, disease, drug
ora_kegg = function(genes, organism = "hsa", db = "pathway", ...) {
    lt = load_kegg_genesets(organism, db)

    df = ora(genes, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}


load_msigdb_genesets = function(collection, version) {
    dataset = paste0(version, "/", collection)
    if(is.null(CACHE$msigdb_genesets[[dataset]])) {
        gs = get_msigdb(version = version, collection = collection)
        CACHE$msigdb_genesets[[dataset]] = gs
    } else {
        gs = CACHE$msigdb_genesets[[dataset]]
    }

    list(gs = gs)
}

ora_msigdb = function(genes, collection = "h.all", version = "2024.1.Hs", ...) {
    lt = load_msigdb_genesets(collection, version)

    df = ora(genes, lt$gs, ...)
    df
}

library(reactome.db)
load_reactome_genesets = function(organism) {
    if(is.null(CACHE$reactome_pathways)) {
        pathway_names = unlist(as.list(reactomePATHID2NAME))
        pathway_gs = as.list(reactomePATHID2EXTID)

        cn = intersect(names(pathway_names), names(pathway_gs))
        pathway_names = pathway_names[cn]
        pathway_gs = pathway_gs[cn]

        CACHE$reactome_pathways$pathway_gs = pathway_gs
        CACHE$reactome_pathways$pathway_names = pathway_names
    } else {
        pathway_gs = CACHE$reactome_pathways$pathway_gs
        pathway_names = CACHE$reactome_pathways$pathway_names
    }

    l = grepl(paste0("^R-", organism, "-"), names(pathway_names))
    list(names = pathway_names[l], gs = pathway_gs[l])
}

ora_reactome = function(genes, organism = "HSA", ...) {
    lt = load_reactome_genesets(organism)

    df = ora(genes, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}

library(UniProtKeywords)
ora_keywords = function(genes, organism = "human", ...) {
    gs = load_keyword_genesets(organism)
    df = ora(genes, gs, ...)
    df
}

library(simona)
load_simona_genesets = function(fun, ontology, organism = "human", ...) {
    if(is.null(CACHE$simona_genesets[[ontology]][[organism]])) {
        dag = fun(organism)
        mc = mcols(dag)
        gs_names = structure(mc$name, names = rownames(mc))
        all_terms = dag_all_terms(dag)
        gs = term_annotations(dag, all_terms)
        gs_names = gs_names[all_terms]

        org_db = switch(organism,
            human = org.Hs.eg.db::org.Hs.eg.db,
            mouse = org.Ms.eg.db::org.Ms.eg.db,
            rat   = org.Rn.eg.db::org.Rn.eg.db,
            stop("not supported")
        )

        x = keys(org_db, keytype = "SYMBOL")
        suppressMessages(map_tb <- AnnotationDbi::select(org_db, keys = x, keytype = "SYMBOL", columns = c("ENTREZID", "GENETYPE")))
        map_tb = map_tb[map_tb$GENETYPE == "protein-coding", , drop = FALSE]
        map_tb = map_tb[!duplicated(map_tb[, 1]), , drop = FALSE]
        map = structure(map_tb[, 2], names = map_tb[, 1])

        gs = lapply(gs, function(x) {
            x2 = map[x]
            unique(x2[!is.na(x2)])
        })

        n = sapply(gs, length)
        gs = gs[n > 0]
        gs_names = gs_names[n > 0]

        CACHE$simona_genesets[[ontology]][[organism]]$gs = gs
        CACHE$simona_genesets[[ontology]][[organism]]$gs_names = gs_names
    } else {
        gs = CACHE$simona_genesets[[ontology]][[organism]]$gs
        gs_names = CACHE$simona_genesets[[ontology]][[organism]]$gs_names
    }

    list(gs = gs, names = gs_names)
}

# the human phenotype ontology
ora_phenotype = function(genes, organism = "human", ...) {
    lt = load_simona_genesets(ontology_hp, "hp", organism = organism)

    df = ora(genes, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}

# disease ontology
ora_disease = function(genes, organism = "human", ...) {
    lt = load_simona_genesets(ontology_rdo, "rdo", organism = organism)

    df = ora(genes, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}
```

Let's test these `ora_*()` functions.

```{r}
genes = random_genes(org.Hs.eg.db, 1000)
head(ora_go(genes))
head(ora_kegg(genes))
head(ora_msigdb(genes))
head(ora_reactome(genes))
head(ora_keywords(genes))
head(ora_phenotype(genes))
head(ora_disease(genes))
```

