---
title: "Topic 1-06: GO/KEGG gene sets for other organisms"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-07: Get GO/KEGG gene sets for other organisms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
rm(list = ls())
```


Besides those well-studied organisms, there are also resources for less well-studies organisms, mainly for KEGG and GO gene sets.


## KEGG pathways

KEGG pathway supports a huge number of organisms. You just need to find the corresponding organism code:

```{r, eval = FALSE}
read.table(url("https://rest.kegg.jp/list/pathway/hsa"), sep = "\t")
read.table(url("https://rest.kegg.jp/list/pathway/ptr"), sep = "\t")
read.table(url("https://rest.kegg.jp/list/pathway/pps"), sep = "\t")
read.table(url("https://rest.kegg.jp/list/pathway/ggo"), sep = "\t")
read.table(url("https://rest.kegg.jp/list/pathway/pon"), sep = "\t")
...
```

Or use the helper function from **GSEAtopics**:

```r
load_kegg_genesets(organism)
```

## OrgDb packages

Biocondutor core team maintaines **org.\*.db** packages for 18 organisms.

```{r, echo = FALSE}
tb = read.table(textConnection(
"org.Hs.eg.db    Human   org.Ss.eg.db    Pig
org.Mm.eg.db    Mouse   org.Gg.eg.db    Chicken
org.Rn.eg.db    Rat org.Mmu.eg.db   Rhesus_monkey
org.Dm.eg.db    Fruit_fly   org.Cf.eg.db    Canine
org.At.tair.db  Arabidopsis org.EcK12.eg.db E_coli_strain_K12
org.Sc.sgd.db   Yeast   org.Xl.eg.db    African_clawed_frog
org.Dr.eg.db    Zebrafish   org.Ag.eg.db    Malaria_mosquito
org.Ce.eg.db    Nematode    org.Pt.eg.db    Chimpanzee
org.Bt.eg.db    Bovine  org.EcSakai.eg.db   E_coli_strain_Sakai
"))
```

```{r, echo = FALSE}
tb[, 1] = paste0("`", tb[, 1], "`")
tb[, 3] = paste0("`", tb[, 3], "`")
tb[, 2] = gsub("_", " ", tb[, 2])
tb[, 4] = gsub("_", " ", tb[, 4])
knitr::kable(tb, col.names = c("Package", "Organism", "Package", "Organism"))
```

These **org.\*.db** packages can be used in the same way as what we have seen with the **org.Hs.eg.db** package.

```r
select(org.*.db, keys = keys(org.*.db), columns = c("GOALL", "ONTOLOGYALL"))
select(org.*.db, keys = "BP", keytype = "ONTOLOGYALL", columns = c("ENTREZID", "GOALL"))
as.list(org.*.egGO2ALLEGS)
```

Note if the primary ID type is not EntreZ ID, the name of the object might be different.

## AnnotationHub

On **AnnotationHub**, there are `OrgDb` objects for a huge number of other organisms (~2000) which can be used
for getting GO genes and mappings between gene IDs.

```{r}
library(AnnotationHub)
ah = AnnotationHub()
```

To search for an organism, using its latin name is more suggested. Also add the "OrgDb" (data class) keyword:

```{r}
query(ah, c("cat", "OrgDb"))
```

```{r}
query(ah, c("Felis catus", "OrgDb"))
```


Besides using `query()` to search for the AnnotationHub dataset, you can
also use the **BiocHubsShiny** package to interactively searching for datasets, or the `ah_shiny()` function from **GSEAtopics**.

Now we download the data.

```{r}
# It is annoying that ID changes between different package versions
org_db = ah[["AH117537"]]  # using `[[` downloads the dataset
```


`org_db` is an `OrgDb` object but contains less information than the **org.\*.db** packages:

```{r}
org_db
columns(org_db)
```

You can obtain the GO gene sets manually by `select()`, taking the `"ENTREZID"` and `"GOALL"` columns:

```{r}
all_genes = keys(org_db, keytype = "ENTREZID")
tb = select(org_db, keys = all_genes, keytype = "ENTREZID", 
    columns = c("GOALL", "ONTOLOGYALL"))
head(tb)
```

You may need to clean the mapping table:

```{r}
tb = tb[!is.na(tb$GOALL), ]
tb = unique(tb)
```



With an `OrgDb` object, you can also use the helper function `load_go_genesets()` in **GSEAtopics** package:

```{r}
library(GSEAtopics)
lt = load_go_genesets(org_db, "BP")
```



## Practice 

### Practice 1 {.tabset}

#### Practice

Try to obtain GO (BP) gene sets for dolphin (latin name: Tursiops truncatus) from  **AnnotationHub**,
and KEGG pathways for giant panda (latin name: Ailuropoda melanoleuca)

#### Solution

```{r}
qu = query(ah, c("Tursiops truncatus", "OrgDb"))
orgdb = ah[[ qu$ah_id ]]
gs = load_go_genesets(orgdb, "BP")
```


```{r}
lt = load_kegg_genesets("aml")
```
