---
title: "Topic 6-01: Visualization"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 6-01: Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## ORA result

### Single ORA result

```{r}
library(GSEAtutorial)

lt = readRDS(system.file("extdata", "ora.rds", package = "GSEAtutorial"))
diff_gene = lt$diff_gene
diff_gene = convert_to_entrez_id(diff_gene)
```

```{r}
library(org.Hs.eg.db)
tb = ora_go(diff_gene, org.Hs.eg.db, "BP")
```

Let's start with the enrichment table from the ORA analysis. 


The mostly-used type of plot is to use simple statistical graphics for a small set of pre-selected gene sets.

```{r}
par(mar = c(4.1, 20, 4, 1))
barplot(tb$n_hits[1:10], horiz = TRUE, names.arg = tb$description[1:10], las = 1)
```

Using **ggplot2** is a better idea for visualizing data.

```{r}
library(ggplot2)
ggplot(tb[1:10, ], aes(x = n_hits, y = description)) + 
    geom_col()
```

Number of DE genes in gene set may not be a good statistic, more commonly used statistics
are log2 fold enrichment or -log10 p.adjust.


```{r}
ggplot(tb[1:10, ], aes(x = log2fe, y = description)) + 
    geom_col()
```

It is also common to add the p-values/adjusted p-values to the bars.

```{r}
ggplot(tb[1:10, ], aes(x = log2fe, y = description)) + 
    geom_col() +
    geom_text(aes(x = log2fe, label = sprintf('%1.e', p_adjust)))
```

By default, **ggplot2** reorders labels alphabetically. You can set the name
as a factor and specify the order there. 

```{r}
ggplot(tb[1:10, ], aes(x = log2fe, y = factor(description, levels = description))) + 
    geom_col()
```

Points are also very frequently used.

```{r}
ggplot(tb[1:10, ], aes(x = log2fe, y = description)) + 
    geom_point() +
    xlim(0, max(tb$log2fe[1:10]))
```

Using dot plot, we can map a second statistic to the size of dots.

```{r}
ggplot(tb[1:10, ], aes(x = log2fe, y = description, size = n_hits)) + 
    geom_point() +
    xlim(0, max(tb$log2fe[1:10]))
```

Even more, we can map a third statistic to the dot colors.

```{r}
ggplot(tb[1:10, ], aes(x = log2fe, y = description, size = n_hits, col = n_hits/n_gs)) + 
    geom_point()

```


A volcano plot which is -log10(p.adjust) vs log2 fold enrichment:

```{r}
ggplot(tb, aes(x = log2fe, y = -log10(p_adjust))) +
    geom_point(col = ifelse(tb$log2fe > 1 & tb$p_adjust < 0.001, "black", "grey")) +
    geom_hline(yintercept = 3, lty = 2) + geom_vline(xintercept = 1, lty = 2)
```

With the volcano plot, we can easily see the preference of enrichment to the sizes of gene sets.

```{r}
ggplot(tb, aes(x = log2fe, y = -log10(p_adjust), color = n_gs, size = n_hits)) +
    geom_point() + scale_colour_distiller(palette = "Spectral")

```


### Multiple ORA results

To visualize multiple ORA enrichment tables in one plot, we need to first prepare a data frame
which combines results for a pre-selected gene sets.

```{r}
tbl = readRDS(system.file("extdata", "lt_enrichment_tables.rds", package = "GSEAtutorial"))

set.seed(666)
terms = sample(tbl[[1]]$gene_set, 10)

tb = NULL
for(nm in names(tbl)) {
    x = tbl[[nm]]
    x = x[x$gene_set %in% terms, colnames(x) != "geneID"]
    x$sample = nm
    tb = rbind(tb, x)
}
```

```{r}
ggplot(tb, aes(x = sample, y = description, size = -log10(p_adjust))) +
    geom_point(color = ifelse(tb$p_adjust < 0.05, "black", "grey"))
```

```{r}
ggplot(tb, aes(x = sample, y = wrap_text(description), size = -log10(p_adjust))) +
    geom_point(color = ifelse(tb$p_adjust < 0.05, "black", "grey"))
```

## GSEA result

### Single GSEA result

```{r}
gene_diff_score = readRDS(system.file("extdata", "gene_diff_score.rds", package = "GSEAtutorial"))
gene_diff_score = convert_to_entrez_id(gene_diff_score)
gene_diff_score = sort(gene_diff_score, decreasing = TRUE)
tb_gsea = fgsea_go(gene_diff_score, org.Hs.eg.db, "BP")
```

Similarly, we can visualize NES scores of gene sets.

```{r}
ggplot(tb_gsea[1:10, ], aes(x = nes, y = description)) + 
    geom_col()
ggplot(tb_gsea[1:10, ], aes(x = nes, y = description)) + 
    geom_col(fill = ifelse(tb_gsea$nes[1:10] > 0, "red", "darkgreen"))
```

The volcano plot:

```{r}
ggplot(tb_gsea, aes(x = nes, y = -log10(p_adjust))) +
    geom_point(col = ifelse(abs(tb_gsea$nes) > 1 & tb_gsea$p_adjust < 0.05, "black", "grey")) +
    geom_hline(yintercept = -log10(0.05), lty = 2) + geom_vline(xintercept = c(1, -1), lty = 2)
```


The degree of enrichment is also dependent to the gene set sizes.

```{r}
ggplot(tb_gsea, aes(x = nes, y = -log10(p_adjust), color = n_gs)) +
    geom_point() + scale_colour_distiller(palette = "Spectral")
```


And the classic GSEA plot:

```{r}
gs = load_go_genesets(org.Hs.eg.db, "BP")$gs
gsea_plot(gene_diff_score, c("GO:0007186", "GO:0022402", "GO:0050877"), gs)
```



