---
title: "Topic 4-02: Single sample GSEA"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 4-02: GSVA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```

We need an expression matrix as well as a collection of gene sets.

```{r}
lt = readRDS(system.file("extdata", "p53_expr.rds", package = "GSEAtraining"))
expr = lt$expr
condition = lt$condition

ln = strsplit(readLines(system.file("extdata", "c2.symbols.gmt", package = "GSEAtraining")), "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])
```



## GSEA per sample

The implementation is quite straightforward. As GSEA expects a vector of gene scores, expression of all genes in a single sample
can also be thought of as a type of gene scores. What we do is only to make sure such expression is relative and corresponds to "differential expression"
of genes in that single sample. This means we need to do a proper normalization or scaling across samples.

We use the simplest normalization method: z-score normalization:

```{r}
expr_scaled = t(scale(t(expr)))
```

The gene set level score in a sample should capture both up- and down-regulation. The default method in GSEA already can capture it
by taking the maximal deviation of two CDFs for genes in and not in the gene set, and keeping the sign. GSVA also uses another metric,
which is the difference between the maximal positive deviation and the maximal negative deviation. We implement both:


```{r}
es_score = function(s, gene_sets, power = 1, method = c("std", "diff")) {
    n = length(s)
    ngs = length(gene_sets)
    es = numeric(ngs)

    method = match.arg(method)[1]

    for(i in seq_len(ngs)) {
        l_set = names(s) %in% gene_sets[[i]]

        s_set = numeric(n)
        s_set[l_set] = abs(s[l_set])^power
    
        l_other = !l_set

        f1 = cumsum(s_set)/sum(s_set)
        f2 = cumsum(l_other)/sum(l_other)
        m1 = max(f1 - f2)
        m2 = min(f1 - f2)

        if(method == "std") {
            es[i] = max(abs(m1), abs(m2)) * ifelse(abs(m1) > abs(m2), sign(m1), sign(m2))
        } else {
            if(m1 >= 0 && m2 >= 0) {
                es[i] = max(m1, m2)
            } else if(m1 <= 0 && m2 <= 0) {
                es[i] = min(m1, m2)
            } else {
                es[i] = m1 + m2
            }
        }
    }

    es
}
```

Now we can go over each sample and calculate a geneset-sample matrix. Let's do it for both "std" and "diff" methods

```{r}
mat_gs1_std = matrix(nrow = length(gs), ncol = ncol(expr))
rownames(mat_gs1_std) = names(gs)
for(i in 1:ncol(expr)) {
    mat_gs1_std[, i] = es_score(sort(expr_scaled[, i], decreasing = TRUE), gs, method = "std")
}

mat_gs1_diff = matrix(nrow = length(gs), ncol = ncol(expr))
rownames(mat_gs1_diff) = names(gs)
for(i in 1:ncol(expr)) {
    mat_gs1_diff[, i] = es_score(sort(expr_scaled[, i], decreasing = TRUE), gs, method = "diff")
}
```

In GSVA, instead of using the normalized expression values, it uses the ranked values as weights. We also calculate it by two ES methods:

```{r}
mat_gs2_std = matrix(nrow = length(gs), ncol = ncol(expr))
rownames(mat_gs2_std) = names(gs)
for(i in 1:ncol(expr)) {
    v = rank(expr_scaled[, i]) - nrow(expr_scaled)/2
    mat_gs2_std[, i] = es_score(sort(v, decreasing = TRUE), gs, method = "std")
}

mat_gs2_diff = matrix(nrow = length(gs), ncol = ncol(expr))
rownames(mat_gs2_diff) = names(gs)
for(i in 1:ncol(expr)) {
    v = rank(expr_scaled[, i]) - nrow(expr_scaled)/2
    mat_gs2_diff[, i] = es_score(sort(v, decreasing = TRUE), gs, method = "diff")
}
```

Let's compare the four matrices:

```{r, fig.width = 8, fig.height = 8}
par(mfrow = c(2, 2))
hist(mat_gs1_std)
hist(mat_gs1_diff)
hist(mat_gs2_std)
hist(mat_gs2_diff)
```

```{r, fig.height = 8, fig.height = 8}
library(ComplexHeatmap)
rod = hclust(dist(mat_gs1_std))$order
cod = hclust(dist(t(mat_gs1_std)))$order

p1 = grid.grabExpr(draw(Heatmap(mat_gs1_std, row_order = rod, column_order = cod, show_row_names = FALSE, column_title = "mat_gs1_std")))
p2 = grid.grabExpr(draw(Heatmap(mat_gs1_diff, row_order = rod, column_order = cod, show_row_names = FALSE, column_title = "mat_gs1_diff")))
p3 = grid.grabExpr(draw(Heatmap(mat_gs2_std, row_order = rod, column_order = cod, show_row_names = FALSE, column_title = "mat_gs2_std")))
p4 = grid.grabExpr(draw(Heatmap(mat_gs2_diff, row_order = rod, column_order = cod, show_row_names = FALSE, column_title = "mat_gs2_diff")))

library(cowplot)
plot_grid(p1, p2, p3, p4, nrow = 2)
```


## GSVA

Next we run the **GSVA** package. 

```{r}
library(GSVA)
mat_gs_gsva = gsva(gsvaParam(log(expr), gs), verbose = FALSE)
```

You can do a log-transformation, this is optional. **GSVA** uses CDF to normalize expression values.


```{r, fig.height = 4, fig.width = 8}
p4 = grid.grabExpr(draw(Heatmap(mat_gs2_diff, row_order = rod, column_order = cod, show_row_names = FALSE, column_title = "mat_gs2_diff")))
p5 = grid.grabExpr(draw(Heatmap(mat_gs_gsva, row_order = rod, column_order = cod, show_row_names = FALSE, show_column_names = FALSE, column_title = "GSVA")))
plot_grid(p4, p5, nrow = 1)
```

## What to do with the geneset-sample matrix

The geneset-sample matrix is the end of the analysis. It only serves as an intermediate step. You still need
to apply, e.g. t-test, to get significant gene sets.


```{r, message=FALSE}
library(genefilter)
# high stat means high in MUT
tdf = rowttests(mat_gs2_std, factor(condition, levels = c("MUT", "WT")))
tdf$fdr = p.adjust(tdf$p.value, "BH")
tdf = tdf[order(tdf$p.value), ]
```

How many gene sets are significant?

```{r}
tdf[tdf$fdr < 0.05, ]
```

## Compare to normal GSEA

The next natural question is, compared to normal GSEA analysis, does GSVA improves the analysis? or how consistent is
between the two types of analysis?

The `tdf` object already contains differential analysis on the geneset-sample matrix and it gives significant gene sets.
For the normal GSEA analysis, we use the same setting as before.


```{r}
library(matrixStats)
m1 = expr[, condition == "WT"]  # only samples in group 1
m2 = expr[, condition == "MUT"]  # only samples in group 2
s = (rowMeans(m2) - rowMeans(m1))/(rowSds(m1) + rowSds(m2)) # gene-level difference scores
s = sort(s, decreasing = TRUE)

tb_gsea = fgsea_wrapper(s, gs)
```


Most significant gene sets from the two methods:

```{r}
head(tdf)
head(tb_gsea)
```


And top 50 gene sets from the two methods:

```{r}
library(eulerr)
plot(euler(list(GSVA = rownames(tdf)[order(tdf$p.value)[1:50]], 
                GSEA = tb_gsea$gene_set[order(tb_gsea$p_value)[1:50]])
          ), quantities = TRUE)
```

```{r, fig.width = 8, fig.height = 4}
rownames(tb_gsea) = tb_gsea$gene_set
cn = intersect(rownames(tdf), rownames(tb_gsea))
par(mfrow = c(1, 2))
plot(tdf[cn, "p.value"], tb_gsea[cn, "p_value"])
plot(tdf[cn, "statistic"], tb_gsea[cn, "nes"])
abline(h = 0, lty = 2, col = "grey")
abline(v = 0, lty = 2, col = "grey")
```

This makes a problem here that one method generates a significant gene set
which can be completely insignificant under another method (left plot). The
direction of the general differentiation of a gene set can be reversed in
different methods.

```{r}
g = intersect(gs[["cell_proliferation"]], rownames(expr))
mm = expr_scaled[g, ]

Heatmap(mm, name = "z-score", top_annotation = HeatmapAnnotation(cond = condition),
    right_annotation = rowAnnotation(bar = anno_barplot(s[g], width = unit(2, "cm"))),
    column_split = condition) %v%
Heatmap(mat_gs_gsva["cell_proliferation", , drop = FALSE], name = "set variation", height = unit(5, "mm")) %v%
Heatmap(mat_gs1_diff["cell_proliferation", , drop = FALSE], name = "set variation", height = unit(5, "mm"))
```


Gene expression might not be a proper ...

## Compare to gene-based analysis

Can geneset-sample matrix reduce the noise from the original gene-sample matrix?


```{r, fig.height = 6}
library(cola)
res1 = consensus_partition(mat_gs_gsva, top_value_method = "ATC", partition_method = "skmeans", verbose = FALSE)
get_signatures(res1, k = suggest_best_k(res1))

res2 = consensus_partition(expr, top_value_method = "ATC", partition_method = "skmeans", verbose = FALSE)
get_signatures(res2, k = suggest_best_k(res2))
```


## Discussion

In what scenario GSVA is useful?

```{r}
Heatmap(mat_gs_gsva[tdf$p.value < 0.01, ])
```

How good it is for subgroup

pathway activity should be an intrinsic attribute of xxx, but GSEA-like actually compares to others
depnence 

why not simply avg(s)