---
title: "Topic 2-05: fgsea"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-05: fgsea}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```

The permutation-based method is by nature very slow, especially if you want to get p-values
in high precision. In this document, we demonstrate the use of an extremely fast package **fgsea**,
but only for the gene-permutation-based method.

We still use the p53 dataset.

```{r}
library(GSEAtopics)
data(p53_dataset)
s = p53_dataset$s
gs = p53_dataset$gs
```

The use of **fgsea** is striaghtforward. Just use the `fgsea()` function:

```{r}
library(fgsea)
df = fgsea(gs, s)
head(df)
```

You only need to make gene IDs in `s` (the name of the vector) are the same as in `gs`.


## Implement helper functions

The same as the `ora_*()` functions for different types of gene sets, we can also implement the `fgsea_*()` functions.

We first define a wrapper function where we rename the colnames of the data frame from `fgsea()`, just let it be
consistent to the results from `ora_*()` functions. We also reorder the result data frame by p-values.

```{r}
fgsea_wrapper = function(s, gs, min_size = 5, max_size = 1000, ...) {
    s = sort(s, decreasing = TRUE)
    df = fgsea(gs, s, minSize = min_size, maxSize = max_size, ...)
    df = as.data.frame(df)
    colnames(df) = c("gene_set", "p_value", "p_adjust", "log2_p_err", "es", "nes", "n_gs", "leading_edge")
    df = df[order(df$p_value), , drop = FALSE]
    rownames(df) = NULL
    df
}
```

Now we can write these helper functions. The logic is the same as `ora_*()` functions.
We first retrieve gene sets than apply `fgsea_wrapper()`.


```{r}
fgsea_go = function(s, org_db = org.Hs.eg.db::org.Hs.eg.db, ontology = "BP") {
    tb = AnnotationDbi::select(org_db, keys = ontology, keytype = "ONTOLOGYALL", 
        columns = c("ENTREZID", "GOALL"))
    gs = split(tb$ENTREZID, tb$GOALL)
    gs = lapply(gs, function(x) {
        as.character(unique(x))
    })
   
    fgsea_wrapper(s, gs)
}

fgsea_kegg = function(s, organism = "hsa") {
    pathway_genes = read.table(url(paste0("https://rest.kegg.jp/link/pathway/", organism)), 
        sep = "\t", quote = "")
    pathway_genes[, 1] = gsub("^\\w+:", "", pathway_genes[, 1])
    pathway_genes[, 2] = gsub("^\\w+:", "", pathway_genes[, 2])
    pathway_genes = split(pathway_genes[, 1], pathway_genes[, 2])

    fgsea_wrapper(s, pathway_genes)
}
```

All these `fgsea_*()` functions use EntreZ IDs as the internal IDs. Then for the input
gene score vector, genes should also have EntreZ IDs.

### Convert gene IDs

We have demonstrated how to convert gene IDs for a vector of genes. Here the scenario is 
a little bit different. Gene IDs are names of a numeric vevtor.

We first generate a global mapping between symbols and EntreZ IDs:

```{r}
library(org.Hs.eg.db)
map = map_to_entrez_id("SYMBOL", org.Hs.eg.db)
```

`map_to_entrez_id()` only returns mapping for protein-coding genes. According to the pratice we have made, the human, the mapping
between symbols and EntreZ IDs is one-to-one. But we need to be careful. 1. gene symbols change from time to time. You cannot ensure
the symbol used in the dataset is the most recent version. 2. for other organisms, we don't know whether the mapping is also one-to-one.
Let's do it in a conservative way.

```{r}
symbol = names(s)
entrez = map[symbol]
```

in `entrez`, there are possible `NA` values or duplicated values. We first remove `NA`s:


```{r}
s2 = unname(s) # since s2 will have EntreZ IDs as names 
l = !is.na(entrez)
s2 = s2[l]
entrez = entrez[l]
```

Then remove duplicates, either only keep the first one, or take the mean.

```r
tapply(s2, entrez, function(x) x[1])
tapply(s2, entrez, function(x) mean(x))
```

I use the mean:

```{r}
s2 = tapply(s2, entrez, mean)
```

In **GSEAtopics**, there is a helper function `convert_to_entrez_id()`
which works on gene ID vecotr, numeric vector with gene IDs as names, and matrix with gene IDs as row names.

```{r}
s2 = convert_to_entrez_id(s) # if is human, the the second argument can be omitted
```

And let's try these `fgsea_*()` helper functions:

```{r}
head(fgsea_go(s2))
head(fgsea_kegg(s2))
```

In **GSEAtopics** there are the following `fgsea_*()` functions:

- `fgsea_go()`
- `fgsea_kegg()`
- `fgsea_msigdb()`
- `fgsea_reactome()`
- `fgsea_keywords()`
- `fgsea_phenotype()`
- `fgsea_disease()`

## Compare to the native implementations

Let's have a quick comparison between the `fgsea()` function and gene-permutation-based 
function we have implemented in the previous document. We use the `gsea_gene_perm()` functionn
from the **GSEAtopics** package.
Under the default setting, both functions use the same statistic and perform two-sided test:


```{r}
df1 = fgsea_wrapper(s, gs)
df2 = gsea_gene_perm(s, gs, verbose = FALSE)
```

We compare p-values and NES from the functions:

```{r, fig.width = 10, fig.height = 5}
rownames(df1) = df1$gene_set
rownames(df2) = df2$gene_set

cn = intersect(rownames(df1), rownames(df2))
par(mfrow = c(1, 2))
plot(df1[cn, "p_value"], df2[cn, "p_value"], main = "p_value", xlab = "fgsea", ylab = "native")
plot(df1[cn, "nes"], df2[cn, "nes"], main = "nes", xlab = "fgsea", ylab = "native")
```

And for the one-sided test which only look for enrichment of up-regulated genes:

```{r, fig.width = 10, fig.height = 5}
df1 = fgsea_wrapper(s, gs, scoreType = "pos")
df2 = gsea_gene_perm(s, gs, direction = "pos", verbose = FALSE)

rownames(df1) = df1$gene_set
rownames(df2) = df2$gene_set

cn = intersect(rownames(df1), rownames(df2))
par(mfrow = c(1, 2))
plot(df1[cn, "p_value"], df2[cn, "p_value"], main = "p_value", xlab = "fgsea", ylab = "native")
plot(df1[cn, "nes"], df2[cn, "nes"], main = "nes", xlab = "fgsea", ylab = "native")
```

## Other ways to capture bi-directional changes

The default behavior of two-sided `fgsea()` or the GSEA method is to capture the stronger one of the up- and down-regulation of genes in a gene set.
There are two other ways:

1. If you want to keep enrichment for both up-regulated and down-regulated genes. You can perform two GSEA analysis, a one-sided test only for up-regulated genes and a second one-sided test only for down-regulated genes.

```r
df1 = fgsea_*(s, gs, scoreType = "pos")
df2 = fgsea_*(s, gs, scoreType = "neg")
```

2. if the gene score is log2fc-like, i.e. with both positive and negative values, and if you do not care the direction of the differential expression,
you can use `|s|` as the gene scores.

```r
fgsea_*(sort(abs(s), decreasing = TRUE), gs, scoreType = "pos")
```

## Which type of gene scores to use

GSEA requires a pre-calculated vector of gene-level scores. In the previous example, we use the signal-to-noise ratio.
As there are many other ways to calculate a gene-level score. We ask, do they all perform similarly or which ones are better
and which ones are worse?

Let's go back to the original dataset, the matrix and the condition vector:

```{r}
expr = p53_dataset$expr
condition = p53_dataset$condition
```

We try the following five metrics: log2 fold change, signal-to-noise ratio, t-value, -log p-value multiplies by the sign of t-values,
SAM moderated t-value.

```{r}
m1 = expr[, condition == "MUT"]
m2 = expr[, condition == "WT"]

library(matrixStats)
log2fc = log2(rowMeans(m1)/rowMeans(m2))
s2n = (rowMeans(m1) - rowMeans(m2))/(rowSds(m1) + rowSds(m2))

library(genefilter)
tdf = rowttests(expr, condition)  # equal variance
t_value = tdf$statistic
names(t_value) = rownames(tdf)
log_p = -log10(tdf$p.value)*sign(tdf$statistic)
names(log_p) = rownames(tdf)

n1 = ncol(m1)
n2 = ncol(m2)
v1 = rowVars(m1)
v2 = rowVars(m2)
sp = sqrt( ((n1-1)*v1 + (n2-1)*v2)/(n1 + n2 - 2) )*sqrt(1/n1 + 1/n2)
sam = (rowMeans(m1) - rowMeans(m2))/(sp + quantile(sp, 0.1))
```

First we can make a pairwise scatter plot to see the similarities of the five metrics:

```{r, fig.width = 8, fig.height = 8}
pairs(data.frame(log2fc = log2fc, s2n = s2n, t_value = t_value, log_p = log_p, sam = sam), pch = ".")
```

We also look at the distribution of the five metrics:

```{r, fig.width = 8, fig.height = 6}
par(mfrow = c(2, 3))
plot(sort(log2fc, decreasing = TRUE), main = "log2fc")
plot(sort(s2n, decreasing = TRUE), main = "s2n")
plot(sort(t_value, decreasing = TRUE), main = "t_value")
plot(sort(log_p, decreasing = TRUE), main = "log_p")
plot(sort(sam, decreasing = TRUE), main = "sam")
```

And we perform GSEA with five gene score vectors.

```{r}
tb_log2fc  = fgsea_wrapper(sort(log2fc, decreasing = TRUE), gs)
tb_s2n     = fgsea_wrapper(sort(s2n, decreasing = TRUE), gs)
tb_t_value = fgsea_wrapper(sort(t_value, decreasing = TRUE), gs)
tb_log_p   = fgsea_wrapper(sort(log_p, decreasing = TRUE), gs)
tb_sam     = fgsea_wrapper(sort(sam, decreasing = TRUE), gs)
```

Top 50 most significant gene sets:

```{r}
library(eulerr)
plot(euler(list(log2fc = tb_log2fc$gene_set[1:50],
                s2n = tb_s2n$gene_set[1:50],
                t_value = tb_t_value$gene_set[1:50],
                log_p = tb_log_p$gene_set[1:50],
                sam = tb_sam$gene_set[1:50])),
    quantities = TRUE)
```

**Conclusion:** use t-value or similar metrics as gene-level scores for GSEA analysis.

