---
title: "Topic 1-08: Gene ID conversion"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-09: Gene ID conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```

Gene ID type conversion is a very common task in gene set enrichment analysis.
We take **org.Hs.eg.db** (for human) as an example.

```{r}
library(org.Hs.eg.db)
```

Note: it is the same for the `OrgDb` objects for other orgainsms on **AnnotationHub**.

## Use the select() interface

We need the following three types of information:

- `keys`: Gene IDs in one ID type;
- `keytypes`: The name of the input ID type;
- `columns`: The name of the output ID type;

To get the valid name of ID types:

```{r}
keytypes(org.Hs.eg.db)
columns(org.Hs.eg.db)
```

### 1:1 mapping

For example, we want to convert the following two genes into Entrez IDs.

```{r}
genes = c("TP53", "MDM2")
```

1. use `select()`. The following function call can be read as "select 'ENTREZID' for the genes where their 'SYMBOL' are in 'gene'".

```{r}
map = select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = "ENTREZID")
map
```

Then we can create a named vector for the conversion:

```{r}
map_vec = structure(map$ENTREZID, names = map$SYMBOL)
map_vec[genes]
```

What if we convert them to Ensembl gene IDs:

```{r}
map = select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = "ENSEMBL")
map
```

If you want to map to multiple ID types:

```{r}
select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = c("ENTREZID", "ENSEMBL"))
```

2. use `mapIds()` which is very similar to `select()`.

Note the argument is named `column` instead of `columns`, so you can only map to one gene ID type.


```{r}
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENTREZID")
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENSEMBL")
```


### 1:many mapping

Now there might be some problems if the mapping is not 1:1.

```{r}
genes = c("TP53", "MMD2")
map = select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = "ENTREZID")
map
```

Usually it is hard to pick one unique gene for such 1:mapping case, but we can add an additional
column "GENETYPE" when querying:

```{r}
map = select(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", columns = c("ENTREZID", "GENETYPE"))
map
```

For "MMD2", adding the "GENETYPE" column works because the second hit of it is annotated to "unknown".
We can simply remove it.

```{r}
map[map$GENETYPE == "protein-coding", ]
```

And it is always a good idea to only inlucde protein-coding genes in gene set enrichment analysis.


When you run `mapIds()`, if there are multiple mappings, by default it only selects the first one,
which may cause problem if the first mapping is not protein-coding.

```{r}
mapIds(org.Hs.eg.db, keys = genes, keytype = "SYMBOL", column = "ENTREZID")
```


## Use the pre-generated objects

In **org.\*.db**, there are also pre-generated objects that already contains mapping between EntreZ IDs to a specific gene ID type.

```{r}
ls(envir = asNamespace("org.Hs.eg.db"))
```

The following six objects can be used to convert between major gene ID types:

- `org.Hs.egENSEMBL`: Entrez -> Ensembl
- `org.Hs.egENSEMBL2EG`: Ensembl -> Entrez
- `org.Hs.egREFSEQ`: Entrez -> RefSeq
- `org.Hs.egREFSEQ2EG`: RefSeq -> Entrez
- `org.Hs.egSYMBOL`: Entrez -> Symbol
- `org.Hs.egSYMBOL2EG`: Symbol -> Entrez

```{r}
org.Hs.egSYMBOL
```

If you have a single gene, you can use `[[`:

```{r}
org.Hs.egSYMBOL2EG[["TP53"]]
```


If you have multiple genes, use `[` + `as.list()`:

```{r}
lt = as.list(org.Hs.egSYMBOL2EG)
lt[genes]

# or
as.list(org.Hs.egSYMBOL2EG[genes])
```

But you have to face the same problem of multiple mappings.

## The strorage type of Entrez IDs

This may cause problems silently to your analysis. Let's say, we have the following six genes with their Entrez IDs and symbols.

Entrez IDs are represented as integers. If the corresponding data is imported from text tables, the column of Entrez IDs will
be automatically saved as integers. 

```{r}
eg = c(1, 3, 6, 8, 10, 37)
symbol = c("gene1", "gene2", "gene3", "gene4", "gene5", "gene6")
```

Now, we want to convert from Entrez IDs to gene symbols, we normally generate a "map" variable where EntreZ IDs are names:

```{r}
map = symbol
names(map) = eg
map
```

And we give `map` the Entrez IDs are indices and we expect it returns the corresponding symbols:
However, if Entrez IDs are saved as integers, there will be a problem:

```{r}
unname(map[eg])
```

The problem is there are both character indicies and integer indices for a vector. In the prevoius usage, we expect `eg` is used
as character indices, but in fact they are used as integer indices. We have to explicitely convert `eg` to characters.

```{r} 
unname(map[as.character(eg)])
```

On the other hand, `eg` as integers is only used in the NCBI database, but on our user-side. It is recommended to save them as characters,
because they are more like a "name representation" of genes.

## Conclusion

If the mappig is 1:1, all the methods metioned is OK. When the mapping is 1:many, use `select()` + filtering by the "GENETYPE" column is safer.


In the **GSEAtutorial** package, there is a `map_to_entrez_id()` which generates a global mapping
variable to EntreZ IDs for protein-coding genes.

```{r}
library(GSEAtutorial)
# symbol to EntreZ ID
map = map_to_entrez_id("SYMBOL", org.Hs.eg.db)
head(map)
```

## Practice


### {.tabset}

#### Practice 1

Using `org.Hs.eg.db`, compare the following three types of gene IDs: `SYMBOL`, `ENTREZID` and `ENSEMBL`.
For two types denoted as $T_1$ and $T_2$, i.e. for the conversion $T_1 \rightarrow T_2$, calculated the 
fraction of IDs in $T_1$ that have multiple mappings. Do the comparison for every combination of $T_1$
and $T_2$ and also do it only restricted in protein-coding genes.

#### Solution


```{r}
t = c("SYMBOL", "ENTREZID", "ENSEMBL")
for(i in 1:3) {
    for(j in 1:3) {
        if(i == j) next
        cat("conversion ", t[i], " -> ", t[j], ": ", sep = "")

        all_ids = keys(org.Hs.eg.db, keytype = t[i])
        suppressMessages(map <- select(org.Hs.eg.db, keys = all_ids, keytype = t[i], columns = t[j]))
        tb = table(map[[1]])
        p = sum(tb > 1)/length(tb)
        cat(p, "\n")
    }
}
```

And only restricted in protein-coding genes:

```{r}
for(i in 1:3) {
    for(j in 1:3) {
        if(i == j) next
        cat("conversion ", t[i], " -> ", t[j], ": ", sep = "")

        all_ids = keys(org.Hs.eg.db, keytype = t[i])
        suppressMessages(map <- select(org.Hs.eg.db, keys = all_ids, keytype = t[i], columns = c(t[j], "GENETYPE")))
        map = map[map$GENETYPE == "protein-coding", ]
        tb = table(map[[1]])
        p = sum(tb > 1)/length(tb)
        cat(p, "\n")
    }
}
```

Two reasons for the multiple mapping between SYMBOL/ENTREZ and ENSEMBL

1. Gene alleles are recorded in ENSENBL, but only the main/curated gene is listed on NCBI.
2. New (unnamed) genes which are very similar as gene $G$ are all listed on NCBI, but they are all assigned to the same IDs as $G$ on Ensembl.



