---
title: "Topic 1-04: Reactome gene sets"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-04: Reactome gene sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6,
    fig.height = 6,
    fig.align = "center"
)
```



Reactome is another popular pathway database. It organizes pathways in a
hierarchical manner, which contains pathways and sub pathways or pathway
components. The up-to-date pathway data can be direclty found at
https://reactome.org/download-data (Mapping Files -> Identifier mapping files -> NCBI to All pathways)


## reactome.db

There is a **reactome.db** package on Bioconductor. But note it is not an official Bioconductor data package.

The basic information of the data:

```{r}
library(reactome.db)
reactome.db
```

The "SOURCEDATE" might be the version of the data.

In **reactome.db**, the important objects are 

- `reactomePATHID2EXTID` contains mappings between reactome pathway IDs and gene EntreZ IDs
- `reactomePATHID2NAME` contains pathway names

```{r}
library(reactome.db)
tb = toTable(reactomePATHID2EXTID)
head(tb)
p2n = toTable(reactomePATHID2NAME)
head(p2n)
```

In the previous code, we use the function `toTable()` to retrieve the data as a data frame. You can also try `as.list()` on the two objects and compare the output.

Reactome contains pathway for multiple organisms. In the reactome ID, the second section contains the organism, e.g. in previous output `HSA`. 

```{r, fig.width = 10.5}
sort(table( gsub("^R-(\\w+)-\\d+$", "\\1", p2n[, 1]) ))
barplot(sort(table( gsub("^R-(\\w+)-\\d+$", "\\1", p2n[, 1]) )))
```

## Directly get data from the database

If you are not sure whether **reactome.db** provides the most up-to-date data, we can directly process the original data fro reactome database.

```{r}
download.file("https://reactome.org/download/current/NCBI2Reactome_All_Levels.txt", destfile = "NCBI2Reactome_All_Levels.txt")
tb = read.table("NCBI2Reactome_All_Levels.txt", sep = "\t", comment.char = "")
head(tb)
```

```{r, echo = FALSE}
invisible(file.remove("NCBI2Reactome_All_Levels.txt"))
```

Restrict to human pathways. **Note do not use column 6 for filtering. **

```{r}
tb2 = tb[grepl("HSA", tb[, 2]), ]
```

The gene sets

```{r}
gs = split(tb2[, 1], tb2[, 2])
```

It is better to ensure Entrez IDs are stored as characters and genes are unique

```{r}
gs = lapply(gs, function(x) unique(as.character(x)))
```

Pathway names:

```{r}
p2n = unique(tb[, c(2, 4)])
```

## Use simona to manually construct gene sets

The file we used "NCBI2Reactome_All_Levels.txt" has the description  
"All levels of the pathway hierarchy" from the reactome website. Note pathways are organized in a hierarchical way (an ontology) where 
if a gene is annotated to a pathway, it should be annotated to all its ancestor pathways. For most gene annotations on ontologies (such as GO),
genes are only annotated to the lowest terms in the ontology. We ask, in "NCBI2Reactome_All_Levels.txt", are all genes correctly assigned
to all ancestor pathways?

Let's manually create an ontology object and merge gene annotations.

Read the hierarchical relations of pathways. Note the first column contains parent pathways and the second column contains child pathways:


```{r}
library(simona)
df = read.table("https://reactome.org/download/current/ReactomePathwaysRelation.txt", sep = "\t")
df = df[grepl("HSA", df[, 1]) & grepl("HSA", df[, 2]), ]
head(df)
```

Create an ontology object with **simona** and also add gene annotations.

```{r}
dag = create_ontology_DAG(df[, 1], df[, 2], annotation = gs)
dag
```

`term_annotations()` automatically merges genes for ancestor pathways:

```{r}
gs3 = term_annotations(dag, setdiff(dag_all_terms(dag), "~~all~~"))
gs3 = gs3[sapply(gs3, length) > 0]
```

For this method, **GSEAtopics** has a helper function `load_reactome_genesets()` that wraps the code above:

```r
lt = load_reactome_genesets("HSA")
lt$name  # pathway names
lt$gs    # pathway gene sets
```

## Compare the three methods

Now we have three different ways to get Reactome pathway gene sets, let's compare their consistency.

Let's first rename the three types of gene sets objects.

```{r}
gs_web = gs
gs1 = as.list(reactomePATHID2EXTID)
gs_pkg = gs1[grepl("HSA", names(gs1))]
gs_ontology = gs3
```

The overlap:

```{r}
library(eulerr)
plot(euler(list(gs_pkg = names(gs_pkg), 
                gs_web = names(gs_web),
                gs_ontology = names(gs_ontology))), quantities = TRUE)
```

`gs_web` and `gs_ontology` have the same set of pathways.

```{r}
setequal(names(gs_web), names(gs_ontology))
```

And we compare the numbers of genes in pathways:

```{r, fig.height = 4, fig.width = 8}
cn = intersect(intersect(names(gs_pkg), names(gs_web)), names(gs_ontology))
n1 = sapply(gs_pkg[cn], length)
n2 = sapply(gs_web[cn], length)
n3 = sapply(gs_ontology[cn], length)

par(mfrow = c(1, 2))
plot(n1, n2, log = "xy", xlab = "reactome.db", ylab = "website")
plot(n2, n3, log = "xy", xlab = "website", ylab = "ontology")
```

Conclusion: if you use reactome pathways, use method 2 or 3.

## Practice 

### Practice 1 {.tabset}

#### Practice

Make the distribution of the numbers of genes in Reactome pathways (use human). 

Print the names of pathways with numbers of genes > 2000.

#### Solution

```{r}
library(GSEAtopics)
lt = load_reactome_genesets("HSA")
n_gene = sapply(lt$gs, length)
hist(n_gene)
```

Make the intervals smaller:

```{r}
hist(n_gene, nc = 100)
```

Pathways more than 2000 genes:

```{r}
n_gene[n_gene > 2000]
```

Their names:

```{r}
lt$name[ names(n_gene[n_gene > 2000]) ]
```


If you go to https://reactome.org/PathwayBrowser/, these big pathways correspond to the pathway clusters on the top level.
