---
title: "Topic 2-05: Compare ORA and GSEA"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-05: Compare ORA and GSEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6,
    fig.height = 6,
    fig.align = "center"
)
```

There are ORA and GSEA for gene set analysis. A natural question is 
which one performs better?
We apply ORA and GSEA on the same p53 dataset and compare the results.

```{r}
library(GSEAtopics)
data(p53_dataset)
expr = p53_dataset$expr
condition = p53_dataset$condition
gs = p53_dataset$gs
s = p53_dataset$s2n
```

## ORA

We need to perform t-test or similar DE test to extract DE genes:

We apply t-tests:

```{r}
library(genefilter)
tdf = rowttests(expr, condition)
tdf$fdr = p.adjust(tdf$p.value, "BH")
sum(tdf$fdr < 0.05)
```

It seems there are very few DE genes. We look at the distribution of the t-statistics:

```{r}
plot(sort(tdf$statistic))
```

Instead of setting a cutoff for FDR, we set a cutoff for t-statistics, just to get
enough number of DE genes.

```{r}
sum(abs(tdf$statistic) > 2)
```

```{r}
diff = rownames(expr)[abs(tdf$statistic) > 2]
```

First we perform ORA using the original C2 gene sets. In the following code,
we explicitly set `universe` to the total genes in the experiment, as we want
to make the "universe genes" the same in both ORA and GSEA analysis.

```{r, message = FALSE}
tb_ora = ora(diff, universe = rownames(expr), gs = gs)
```


## GSEA

Next we perform GSEA using signal-to-noise ratios as gene scores. 

```{r}
tb_gsea = fgsea_wrapper(s, gs)
```


## Comparison

We compare the two significant GO lists:

```{r}
library(eulerr)
plot(euler(list(ORA = tb_ora$gene_set[tb_ora$p_value < 0.05], 
                GSEA = tb_gsea$gene_set[tb_gsea$p_value < 0.05])
          ), quantities = TRUE)
```

And we compare the p-values and fold enrichment from the two methods:

```{r, fig.width = 10, fig.height = 5}
rownames(tb_ora) = tb_ora$gene_set
rownames(tb_gsea) = tb_gsea$gene_set
cn = intersect(rownames(tb_ora), rownames(tb_gsea))

par(mfrow = c(1, 2))
plot(tb_ora[cn, "p_value"], tb_gsea[cn, "p_value"])
plot(tb_ora[cn, "log2fe"], log2(abs(tb_gsea[cn, "nes"])))
```

So we would say in general, ORA and GSEA are not comparible as they correspond to two different test answering different questions.

However, they might agree very well for top-significant gene sets:

```{r}
head(tb_ora)
head(tb_gsea)
```

Next let's use a larger gene set collection, the GO BP gene sets. To use the `ora_go()` and `fgsea_go()` helper functions,
we need to convert genes all to EntreZ IDs.

We also do ORA using the default universe and all genes in the experiment as universe. Note the p53 dataset is very old and
the total number of genes in the experiment is quite small.

```{r}
tb_ora1 = ora_go(convert_to_entrez_id(diff))
tb_ora2 = ora_go(convert_to_entrez_id(diff), universe = convert_to_entrez_id(rownames(expr)))
tb_gsea = fgsea_go(convert_to_entrez_id(s))

lt_GO = list(ORA1 = tb_ora1$gene_set[order(tb_ora1$p_value)[1:100]], 
              ORA2 = tb_ora2$gene_set[order(tb_ora2$p_value)[1:100]], 
              GSEA = tb_gsea$gene_set[order(tb_gsea$p_value)[1:100]])
plot(euler(lt_GO), quantities = TRUE)
```

The agreement on individual is so bad, even for two ORA settings.

Although there are big inconsistency between different settings, but note GO has a hierarchical semantic relations,
the following plot shows the general biological functions are very consistent for the three methods.

```{r, fig.width = 10, fig.height = 5}
library(simplifyEnrichment)
simplifyGOFromMultipleLists(lt_GO)
```
