---
title: "Topic 1-02: KEGG gene sets"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-02: KEGG gene sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6,
    fig.height = 6,
    fig.align = "center"
)
```

<style>
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    background: #eee;
    border-left: 6px #CCCCCC solid;
}
</style>

KEGG is a comprehensive database which not only contains pathways but also a large variaty 
of other information. The entrence of getting KEGG data programmatically is via the KEGG REST API.

Please note the restrictions of using KEGG data (https://www.kegg.jp/kegg/legal.html):

> <b>Academic use of KEGG</b><br>
> Academic users may freely use the KEGG website at https://www.kegg.jp/ or its mirror site at GenomeNet https://www.genome.jp/kegg/. Academic users who utilize KEGG for providing academic services are requested to obtain an academic service provider license, which is included in the KEGG FTP academic subscription. The KEGG FTP academic subscription, which is a paid service (see background information), may also be obtained to conveniently download the entire KEGG database.
>
> <b>Non-academic use of KEGG</b><br>
> Non-academic users must understand that KEGG is not a public database and non-academic use of KEGG generally requires a commercial license. There are two types of commercial licenses available: end user and business. The end user license includes access rights to the FTP site and the website, while the business license includes access rights to the FTP site only. Please contact Pathway Solutions for more details.

## Read from the KEGG API

KEGG provides its data via a REST API (https://rest.kegg.jp/). There are
several commands that can be used to retrieve specific types of data. The URL form of the request is 

```
https://rest.kegg.jp/<operation>/<argument>[/<argument2[/<argument3> ...]]
```

To get the KEGG pathway gene sets, we will use the following two operators:

- `link`: get the mapping between two sources.
- `list`: get the details of a list of items.

The `link` operator returns the mapping between two sources of information. We can use
the following command to get the mappings between genes and pathways for human.


```{r}
df1 = read.table(url("https://rest.kegg.jp/link/pathway/hsa"), sep = "\t")
head(df1)
```

In the example, `url()` constructs a connection object that directly transfer
data from the remote URL. You can also first download the output from KEGG into
a local file, read it and later delete the temporary file. `url()` basically automates
such steps.

```{r, eval = FALSE}
temp = tempfile()
download.file("https://rest.kegg.jp/link/pathway/hsa", destfile = temp)
df1 = read.table(temp, sep = "\t")
file.remove(temp)
```

Also you can omit `url()` to directly read from an URL:

```{r, eval = FALSE}
df1 = read.table("https://rest.kegg.jp/link/pathway/hsa", sep = "\t")
```


The URL `"https://rest.kegg.jp/link/hsa/pathway"` returns identical results which
only switches the two columns in the table.

In the output, the first column contains Entrez ID
(users may remove the `"hsa:"` prefix for downstream analysis) and the second
columncontains KEGG pathways IDs (users may remove the `"path:"` prefix).

```{r}
df1[, 1] = gsub("hsa:", "", df1[, 1])
df1[, 2] = gsub("path:",  "", df1[, 2])
head(df1)
```

To get the full name of pathways, use the `list` command:

```{r}
df2 = read.table(url("https://rest.kegg.jp/list/pathway/hsa"), sep = "\t")
head(df2)
df2[, 2] = gsub(" - Homo.*$", "", df2[, 2])
head(df2)
```


## Highlight genes

KEGG also provides pathways as graphs (figures). You may want to highlight differential genes on the graph to see how they specifically affect the pathway.
All you need to do is to prepare a two-column table where the first column contains genes and the second column contains color-settings.

Let's take the pathway "hsa04110" as an example. We color 10 random genes.

```{r, comment = ""}
library(circlize)
df1 = read.table("https://rest.kegg.jp/link/pathway/hsa", sep = "\t")
genes = df1[df1[, 2] == "path:hsa04110", 1]

diff_genes = sample(genes, 10)
settings = data.frame(
    genes = diff_genes, 
    colors = rand_color(10)
)
write.table(settings, file = stdout(), row.names = FALSE, 
    col.names = FALSE, quote = FALSE, sep = "\t")
```

Then open https://www.kegg.jp/pathway/hsa04110.

The value of colors can be a color name (e.g. red, white) or a hexadecimal code (e.g. `#0000FF`)

Note the prefix "hsa:" of gene IDs has no effect. You can keep it or remove it. For the discrete color mapping,
the second column can contain two colors in one row in the form of `color1,color2` which corresponds to the filled
color and the border color. Note the border color also includes the color of the text.

```{r, comment = ""}
settings = data.frame(
    genes = diff_genes, 
    colors = paste0(rand_color(10), ",", sample(colors(), 10))
)
write.table(settings, file = stdout(), row.names = FALSE, 
    col.names = FALSE, quote = FALSE, sep = "\t")
```

To use continuous color mapping, the second column should contain numeric values.

```{r, comment = ""}
settings = data.frame(
    genes = diff_genes, 
    colors = runif(10)
)
write.table(settings, file = stdout(), row.names = FALSE, 
    col.names = FALSE, quote = FALSE, sep = "\t")
```

How to set gene colors: https://www.genome.jp/kegg/webapp/color_gui.html

The package **ggkegg** also supports customizing KEGG graph.

## Practice

### Practice 1 {.tabset}

#### Practice

In the previous example, we obtained the mapping between genes and KEGG pathways for human, using the
organism code of `"hsa"`. Since human is very well-studied and everyone knows `"hsa"` is for human, what if
you are studying dogs? Try to find out the organism code for dog and get the gene-pathway mappings for dog.

Hint: use the `list` operator in KEGG API and search for dog.

#### Solution

You can get organism codes for all organisms by https://rest.kegg.jp/list/organism. Then you
will find the code for dog is `"cfa"`. 


```{r, eval = FALSE}
df = read.table(url("https://rest.kegg.jp/link/pathway/cfa"), sep = "\t")
```


### Practice 2 {.tabset}

#### Practice

Make the distribution of the numbers of genes in KEGG pathways (for human). 

There is a pathway that contains far more genes than all other pathways. What is the pathway ID of it and what is its full name?
Go to the web page of this largest pathway to see its real structure.

#### Solution

```{r}
df1 = read.table(url("https://rest.kegg.jp/link/pathway/hsa"), sep = "\t")
n_gene = table(df1[, 2])
hist(n_gene)
```

Making the intervals smaller is better:

```{r}
hist(n_gene, nc = 100)
```

The largest pathway:

```{r}
which.max(n_gene)
```

Its name:

```{r}
df2 = read.table(url("https://rest.kegg.jp/list/pathway/hsa"), sep = "\t")
df2[df2[, 1] == "hsa01100", ]
```

And its url on KEGG: https://www.kegg.jp/pathway/hsa04110.


### Practice 3 {.tabset}

#### Practice

Highlight the following genes "TGFB1", "TP53", "CDC7" and "E2F1" on the pathway image of "Cell cycle" (organism: human). Choose colors that you like.

Note: you need to first convert gene symbols to Entrez IDs.

#### Solution

```{r}
df2 = read.table(url("https://rest.kegg.jp/list/pathway/hsa"), sep = "\t")
df2[, 2] = gsub(" - Homo .*$", "", df2[, 2])
pathway_id = df2[df2[, 2] == "Cell cycle", 1]
pathway_id
```

Convert symbols to Entrez IDs:

```{r}
library(org.Hs.eg.db)
genes = mapIds(org.Hs.eg.db, keys = c("TGFB1", "TP53", "CDC7", "E2F1"), 
    keytype = "SYMBOL", column = "ENTREZID")
genes
```


Prepare the color table:

```
7040    red,blue
7157    blue,red
8317    yello,darkgreen
1869    grey,red
```

Then go to https://www.genome.jp/pathway/`r pathway_id` and paste the color table there.

