---
title: "Topic 1-04: Get pathways from Reactome"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-04: Get pathways from Reactome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```



Reactome is another popular pathway database. It organizes pathways in a
hierarchical manner, which contains pathways and sub pathways or pathway
components. The up-to-date pathway data can be direclty found at
https://reactome.org/download-data (Mapping Files -> Identifier mapping files -> NCBI to All pathways)


## reactome.db

There is a **reactome.db** on Bioconductor. 

The version of the data:

```{r}
library(reactome.db)
reactome.db
```


In it, the important objects are 

- `reactomePATHID2EXTID` contains mappings between reacotme pathway IDs and gene entrez IDs
- `reactomePATHID2NAME` contains pathway names

```{r}
library(reactome.db)
tb = toTable(reactomePATHID2EXTID)
head(tb)
p2n = toTable(reactomePATHID2NAME)
head(p2n)
```

In the previous code, we use the function `toTable()` to retrieve the data as a data frame. You can also try `as.list()` on the two objects and compare the output.

Reactome contains pathway for multiple organisms. In the reactome ID, teh second section contains the organism, e.g. in previous output `HSA`. 

```{r}
sort(table( gsub("^R-(\\w+)-\\d+$", "\\1", p2n[, 1]) ))
barplot(sort(table( gsub("^R-(\\w+)-\\d+$", "\\1", p2n[, 1]) )))
```

## Directly get data from the database

If you are not sure whether **reactome.db** provides the most up-to-date data, we can directly process the original data fro reactome database.

```{r}
download.file("https://reactome.org/download/current/NCBI2Reactome_All_Levels.txt", destfile = "NCBI2Reactome_All_Levels.txt")
tb = read.table("NCBI2Reactome_All_Levels.txt", sep = "\t", comment.char = "")
head(tb)
```

Restrict to human

```{r}
tb = tb[tb[, 6] == "Homo sapiens", ]
```

The gene sets

```{r}
gs = split(tb[, 1], tb[, 2])
```

It is better to ensure Entrez IDs are stored as characters.

```{r}
gs = lapply(gs, as.character)
```

Pathway names:

```{r}
p2n = unique(tb[, c(2, 4)])
```

## Practice 

### {.tabset}

#### Practice 1

Make the distribution of the numbers of genes in Reactome pathways (use human). 

Print the names of pathways with numbers of genes > 2000.

#### Solution

```{r}
tb = toTable(reactomePATHID2EXTID)
tb = tb[grep("-HSA-", tb[, 1]), ]
n_gene = table(tb[, 1])
hist(n_gene)
```

Making the intervals smaller is better:

```{r}
hist(n_gene, nc = 100)
```

Pathways more than 2000 genes:

```{r}
n_gene[n_gene > 2000]
```

Their names:

```{r}
p2n = toTable(reactomePATHID2NAME)
p2n[p2n[, 1] %in% names(n_gene[n_gene > 2000]), ]
```


If you go to https://reactome.org/PathwayBrowser/, these big pathways correspond to the pathway clusters on the top level.
