---
title: "Topic 1-00: Data representation of gene sets in R"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 1-00: Data representation of gene sets in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```


Gene sets are represented as a list of genes. There are three types of representations in R.


1. A list of gene vectors

```{r}
lt = list(
    geneset1 = c("gene1", "gene2", "gene3"),
    geneset2 = c("gene2", "gene4"),
    geneset3 = c("gene1", "gene3", "gene5", "gene6")
)
lt
```

2. A two-column data frame

```{r}
df = data.frame(
    geneset = c(rep("geneset1", 3), rep("geneset2", 2), rep("geneset3", 4)),
    gene = c("gene1", "gene2", "gene3", "gene2", "gene4", "gene1", "gene3", "gene5", "gene6")
)
df
```

Some tools may need genes to be in the first column:

```{r}
df[, 2:1]
```


These two formats can be very easily converted to each other:

- list to data frame

```{r}
data.frame(
    geneset = rep(names(lt), times = sapply(lt, length)),
    gene = unlist(lt)
)
```

- data frame to list

```{r}
split(df$gene, df$geneset)
```

In the **GSAEtopics** package, there are two helper functions `list_to_data_frame()` and
`data_frame_to_list()` that do the conversions:

```{r}
library(GSEAtopics)
list_to_data_frame(lt)
```

`data_frame_to_list()` automatically guesses which columns are genesets and which columns are genes.

```{r}
data_frame_to_list(df)
```

3. a binary matrix

Not very often, the relation between genes and gene sets can be represented as a binary matrix:

```{r}
m = matrix(0, nrow = 3, ncol = 6)
rownames(m) = unique(df$geneset)
colnames(m) = unique(df$gene)

for(i in seq_len(nrow(df))) {
    m[df[i, 1], df[i, 2]] = 1
}
m
```

It might be more useful when genes have weights:

```{r}
for(i in seq_len(nrow(df))) {
    m[df[i, 1], df[i, 2]] = runif(1)
}
m
```

Note when there are more gene sets, there will be a lot of zeros in `m`. In this case,
we can use "sparse matrix format" to store the data.
