---
title: "Topic 3-02: Local GREAT analysis"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 3-02: Local GREAT analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6,
    fig.height = 6,
    fig.align = "center"
)
```

## Basic usage

First let's load the **rGREAT** package and generate a set of random regions:

```{r}
library(rGREAT)
set.seed(123)
gr = randomRegions(nr = 1000, genome = "hg19")
```


```{r, echo = FALSE}
great_opt$verbose = FALSE
```

Besides the input regions, GREAT analysis needs two types of input data: the gene sets and TSS source. Thus
the second argument in `great()` corresponds to the gene sets, and the
third argument corresponds to the source of TSS.

**rGREAT** has integrated many gene sets and genomes.

The second argument (the gene set) supports:

1. a name of a pre-defined gene set collection
2. a list of genes

The third argument (the source of TSSs) supports

1. genome version,
2. a `GRanges()` object with an additional meta column `gene_id`
3. other not for end users.


```{r}
res = great(gr, "MSigDB:H", "hg19")
```

Following functions are very similar as in online GREAT.

```{r}
tb = getEnrichmentTable(res)
plotVolcano(res)
```

```{r, fig.width = 10}
plotRegionGeneAssociations(res)
getRegionGeneAssociations(res)
plotRegionGeneAssociations(res, term_id = "HALLMARK_APOPTOSIS")
# shinyReport(res)
```

The following gene sets are directly supported:

- `"GO:BP"`: Biological Process.
- `"GO:CC"`: Cellular Component.
- `"GO:MP"`: Molecular Function.

Note MSigDB genes are only for human.

- `"msigdb:H"` Hallmark gene sets.
- `"msigdb:C1"` Positional gene sets.
- `"msigdb:C2"` Curated gene sets.
- `"msigdb:C2:CGP"` C2 subcategory: chemical and genetic perturbations gene sets. 
- `"msigdb:C2:CP"` C2 subcategory: canonical pathways gene sets. 
- `"msigdb:C2:CP:KEGG"` C2 subcategory: KEGG subset of CP.
- `"msigdb:C2:CP:PID"` C2 subcategory: PID subset of CP.
- `"msigdb:C2:CP:REACTOME"` C2 subcategory: REACTOME subset of CP.
- `"msigdb:C2:CP:WIKIPATHWAYS"` C2 subcategory: WIKIPATHWAYS subset of CP.
- `"msigdb:C3"` Regulatory target gene sets.
- `"msigdb:C3:MIR:MIRDB"` miRDB of microRNA targets gene sets.
- `"msigdb:C3:MIR:MIR_LEGACY"` MIR_Legacy of MIRDB.
- `"msigdb:C3:TFT:GTRD"` GTRD transcription factor targets gene sets.
- `"msigdb:C3:TFT:TFT_LEGACY"` TFT_Legacy.
- `"msigdb:C4"` Computational gene sets.
- `"msigdb:C4:CGN"` C4 subcategory: cancer gene neighborhoods gene sets.
- `"msigdb:C4:CM"` C4 subcategory: cancer modules gene sets.
- `"msigdb:C5"` Ontology gene sets.
- `"msigdb:C5:GO:BP"` C5 subcategory: BP subset.
- `"msigdb:C5:GO:CC"` C5 subcategory: CC subset.
- `"msigdb:C5:GO:MF"` C5 subcategory: MF subset.
- `"msigdb:C5:HPO"` C5 subcategory: human phenotype ontology gene sets.
- `"msigdb:C6"` Oncogenic signature gene sets.
- `"msigdb:C7"` Immunologic signature gene sets.
- `"msigdb:C7:IMMUNESIGDB"` ImmuneSigDB subset of C7.
- `"msigdb:C7:VAX"` C7 subcategory: vaccine response gene sets.
- `"msigdb:C8"` Cell type signature gene sets.

The `"GO:"` and "msigdb:" prefix can be removed when specifying in `great()`.

## Setting genome background

Background can be set via `backgroud` and `excluce` arguments.

```r
gap = getGapFromUCSC("hg19", paste0("chr", c(1:22, "X", "Y")))
great(gr, "MSigDB:H", "hg19", exclude = gap)

great(gr, "GO:BP", "hg19", background = paste0("chr", 1:22))
great(gr, "GO:BP", "hg19", exclude = c("chrX", "chrY"))
```

Restricting the anallysis in a certain type of genome background helps to
reduce the false positives. Examples are:

- excluding unmappable regions
- regions with certain level of CG density
- regions with certain level of gene densities
- exons (coding regions)

A very useful case is to perform GREAT only in promoter regions or in distal regions from gene:

```r
great(gr, ..., background = promoters)
great(gr, ..., background = distal)
great(gr, ..., background = gene_body)
```

**Question:** how to obtain intergenic and distal regions from TSS?


## Extend from the complete gene

By default genes are extended only from TSS. In the newer version of **rGREAT** (>= 2.10.0),
you can also extend from the complete gene bodies.

```r
great(gr, "BP", "hg19", extend_from = "gene")
```

## Using other gene sets

`great()` supports self-defined gene sets. Just make sure genes in gene models have
the same ID type as in gene sets.

More gene sets can be set as a list of gene vectors. For most organisms, EntreZ IDs
should be used as the gene ID type in gene sets.

From a `.gmt` file:

```{r}
gs = read_gmt(url("http://dsigdb.tanlab.org/Downloads/D2_LINCS.gmt"), 
    from = "SYMBOL", to = "ENTREZ", orgdb = "org.Hs.eg.db")
gs[1:2]
great(gr, gs, "hg19")
```

KEGG pathways:

```{r}
df = read.table(url("https://rest.kegg.jp/link/pathway/hsa"), sep = "\t")
df[, 1] = gsub("hsa:", "", df[, 1])
df[, 2] = gsub("path:",  "", df[, 2])
gs_kegg = split(df[, 1], df[, 2])
gs_kegg[1:2]
great(gr, gs_kegg, "hg19")
```

## Other organisms which have TxDb packages

On Bioconductor, there are a family of annotation packages with name **TxDb.\*.db**.
These packages contain sources of TSS definitions.

All supported **txdb.\*.db** packages:

```{r}
tb = rGREAT:::BIOC_ANNO_PKGS
knitr::kable(tb[!duplicated(tb$genome_version_in_txdb), ])
```

For these packages, users only need to specify the "genome version":

```r
great(gr, "BP", "mm10")
great(gr, "BP", "rn7")
```

Note `gr` should have `chr` prefix for most of the organisms.


## Use MSigDB gene sets for other organisms

Taking chimpanzee as an example. There is an TxDb on Bioconductor.

```{r}
gr = randomRegions(nr = 1000, genome = "panTro6")
```

```{r}
library(GSEAtopics)
gs = get_msigdb(version = "2024.1.Hs", collection = "h.all")
gs2 = gs_map_to_orthologues(gs, from = "Homo.sapiens", to = "Pan.troglodytes")

great(gr, gs2, "panTro6")
```


## Other organisms who have TxDb object on AnnotationHub

For some other organisms, although they don't have a **TxDb** package on Bioconductor,
but they have a `TxDb` object on **AnnotationHub**:

```{r}
library(AnnotationHub)
ah = AnnotationHub()
query(ah, "TxDb")
```

But it seems many of those organisms are microbes. I will not introduce it here.


## Other organisms who do not have TxDb objects

### Get TSS from NCBI

We need to first know the "accession number" of an organism's genome. Taking
dolphin as an example, its web page on NCBI Genome is https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_011762595.1/. In **rGREAT**, there is a `getGenomeDataFromNCBI()` function which
can automatically retrieve gene definitions from NCBI.

```{r}
genes = getGenomeDataFromNCBI("GCF_011762595.1", return_granges = TRUE)
genes
```

With it, the chromosome lengths are also retrieved:

```{r}
seqlengths(genes)
```

As we have already introduced, retrieve GO gene sets for an organism is easy:

```{r}
library(AnnotationHub)
ah = AnnotationHub()
hit = query(ah, c("Tursiops truncatus", "OrgDb"))
orgdb = ah[[ hit$ah_id ]]

library(GSEAtraining)
gs = get_GO_gene_sets_from_orgdb(orgdb, "BP")
gs[1:2]
```

Note `genes` is set to the argument `tss_source`:

```{r}
gr = randomRegions(nr = 1000, seqlengths =  seqlengths(genes))
great(gr, gs, genes)
```

Remember to double check the format of chromosome names in `gr` and in `genes`.

## Practice

### Practice 1 {.tabset}

#### Practice


Run local GREAT analysis with a list of TFBS from UCSC table browser:

```
clade: Mammal
genome: Human
assembly: GRCh37/hg19
group: Regulation
track: ENCODE 3 TFBS
table: 22Rv1 ZFX (encTfChipRkENCFF445OFX)
```

In the "output format" field, select "BED - browser extensible data", then click the button "get output". In the next page, click the button "get BED".

Use 1. the GO BP gene sets and 2. KEGG pathways.

