---
title: "Topic 2-05: fgsea"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-05: fgsea}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```

The permutation-based method is by nature very slow, especially if you want to get p-values
in high precision. In this document, we demonstrate the use of a extremely efficient package **fgsea**,
but only for the gene-permutation-based method.

We still use the p53 dataset.

```{r}
lt = readRDS(system.file("extdata", "p53_expr.rds", package = "GSEAtutorial"))
expr = lt$expr
condition = lt$condition

ln = strsplit(readLines(system.file("extdata", "c2.symbols.gmt", package = "GSEAtutorial")), "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])

library(matrixStats)
m1 = expr[, condition == "WT"]  # only samples in group 1
m2 = expr[, condition == "MUT"]  # only samples in group 2
s = (rowMeans(m1) - rowMeans(m2))/(rowSds(m1) + rowSds(m2)) # gene-level difference scores
s = sort(s, decreasing = TRUE)
```

The use of **fgsea** is striaghtforward, with the `fgsea()` function:

```{r}
library(fgsea)
df = fgsea(gs, s)
head(df)
```

You only need to make gene IDs in `s` (as the name of the vector) are the same as in `gs`.


## Implement helper functions

The same as the `ora_*()` functions for different types of gene sets, we can also implement the `fgsea_*()` functions.

We first define a wrapper function where we can rename the colnames of the data frame from `fgsea()`, just let it be
consistent to the results from `ora_*()` functions. We also reorder the result data frame by p-values.

```{r}
fgsea_wrapper = function(s, gs, min_size = 5, max_size = 1000, ...) {
    s = sort(s, decreasing = TRUE)
    df = fgsea(gs, s, minSize = min_size, maxSize = max_size, ...)
    df = as.data.frame(df)
    colnames(df) = c("gene_set", "p_value", "p_adjust", "log2_p_err", "es", "nes", "n_gs", "leading_edge")
    df = df[order(df$p_value), , drop = FALSE]
    rownames(df) = NULL
    df
}
```

Now we can write these helper functions. The `load_*_genesets()` functions are already defined in the previous document.


```{r}
fgsea_go = function(s, org_db = org.Hs.eg.db::org.Hs.eg.db, ontology = "BP", ...) {
    lt = load_go_genesets(org_db, ontology)
   
    df = fgsea_wrapper(s, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}

# db: pathway, module, ko, network, disease, drug
fgsea_kegg = function(s, organism = "hsa", db = "pathway", ...) {
    lt = load_kegg_genesets(organism, db)

    df = fgsea_wrapper(s, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}

fgsea_msigdb = function(s, collection = "h.all", version = "2024.1.Hs", ...) {
    lt = load_msigdb_genesets(collection, version)

    df = fgsea_wrapper(s, lt$gs, ...)
    df
}

fgsea_reactome = function(s, organism = "HSA", ...) {
    lt = load_reactome_genesets(organism)

    df = fgsea_wrapper(s, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}

fgsea_keywords = function(s, organism = "human", ...) {
    gs = load_keyword_genesets(organism)
    
    df = fgsea_wrapper(s, gs, ...)
    df
}

# the human phenotype ontology
fgsea_phenotype = function(s, organism = "human", ...) {
    lt = load_simona_genesets(ontology_hp, "hp", organism = organism)

    df = fgsea_wrapper(s, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}

# disease ontology
fgsea_disease = function(s, organism = "human", ...) {
    lt = load_simona_genesets(ontology_rdo, "rdo", organism = organism)

    df = fgsea_wrapper(s, lt$gs, ...)
    df$description = lt$names[df$gene_set]
    df
}
```

Keep in mind, all these `fgsea_*()` functions use EntreZ IDs as the internal IDs. Then for the input
gene score vector, genes should also have EntreZ IDs.


Use the helper function first to generate a global mapping between symbols and EntreZ IDs:

```{r}
library(org.Hs.eg.db)
map = map_to_entrez_id("SYMBOL", org.Hs.eg.db)
```

`map_to_entrez_id()` only returns mapping for protein-coding genes. According to the pratice we have made, the human, the mapping
between symbols and EntreZ IDs is one-to-one. But we need to be careful. 1. gene symbols change from time to time. You cannot ensure
the symbol used in the dataset is the most recent version. 2. for other organisms, we don't know whether the mapping is also one-to-one.
Let's do it in a conservative way.

```{r}
symbol = names(s)
entrez = map[symbol]
```

in `entrez`, there are possible `NA` values or duplicated values. We first remove `NA`s:


```{r}
s2 = unname(s)
s2 = s2[!is.na(entrez)]
entrez = entrez[!is.na(entrez)]
```

Then remove duplicates, either only keep the first one, or take the mean.

```r
tapply(s2, entrez, function(x) x[1])
tapply(s2, entrez, function(x) mean(x))
```

In this exmaple, I would only take the first one

```{r}
s2 = tapply(s2, entrez, function(x) x[1])
```

This is how it is implemented in `convert_to_entrez_id()`.

```{r}
s2 = convert_to_entrez_id(s)
```

And let's try these `fgsea_*()` helper functions:

```{r}
head(fgsea_go(s2))
head(fgsea_kegg(s2))
head(fgsea_msigdb(s2))
head(fgsea_reactome(s2))
head(fgsea_keywords(s2))
head(fgsea_phenotype(s2))
head(fgsea_disease(s2))
```

## Compare to the native implementations

Let's have a quick comparison between the `fgsea()` function and gene-permutation-based function we have implemented in the previous document.
Under the default setting, both functions perform two-sided test:


```{r, fig.width = 10, fig.height = 5}
df1 = fgsea_wrapper(s, gs)
df2 = gsea_gene_perm(s, gs, verbose = FALSE)

rownames(df1) = df1$gene_set
rownames(df2) = df2$gene_set

cn = intersect(rownames(df1), rownames(df2))
par(mfrow = c(1, 2))
plot(df1[cn, "p_value"], df2[cn, "p_value"], main = "p_value", xlab = "fgsea", ylab = "native")
plot(df1[cn, "nes"], df2[cn, "nes"], main = "nes", xlab = "fgsea", ylab = "native")
```

And the one-sided test:

```{r, fig.width = 10, fig.height = 5}
df1 = fgsea_wrapper(s, gs, scoreType = "pos")
df2 = gsea_gene_perm(s, gs, direction = "pos", verbose = FALSE)

rownames(df1) = df1$gene_set
rownames(df2) = df2$gene_set

cn = intersect(rownames(df1), rownames(df2))
par(mfrow = c(1, 2))
plot(df1[cn, "p_value"], df2[cn, "p_value"], main = "p_value", xlab = "fgsea", ylab = "native")
plot(df1[cn, "nes"], df2[cn, "nes"], main = "nes", xlab = "fgsea", ylab = "native")
```

## Other options to capture bi-directional changes

The default behavior of two-sided `fgsea()` or the GSEA method is to capture the stronger one of the up- and down-regulation of genes in a gene set.
There are 

1. if you want to keep enrichment for both up-regulated and down-regulated genes. You can perform two GSEA analysis, a one-sided test only for up-regulated genes and a second one-sided test only for down-regulated genes.

```r
df1 = fgsea_*(s, gs, scoreType = "pos")
df2 = fgsea_*(s, gs, scoreType = "neg")
```

2. if the gene score is log2fc-like, i.e. with both positive and negative values, if you do not care the direction of the differential expression,
you can use `|s|` as the gene scores.

```r
fgsea_*(abs(s), gs, scoreType = "pos")
```

## Which type of gene scores to use

```{r}

```
