---
title: "Topic 4-01: Centrality-based pathway enrichment analysis"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 4-01: Centrality-based pathway enrichment analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 6,
    fig.height = 6,
    fig.align = "center"
)
rm(list = ls())
```

## ORA extension

We start with a differential gene list and an universe gene list. Let's first convert them to EntreZ IDs:

```{r}
library(CePa)
data(gene.list)
genes = gene.list$dif
universe = gene.list$bk
```

```{r}
library(GSEAtopics)
library(org.Hs.eg.db)
genes = convert_to_entrez(genes)
universe = convert_to_entrez(universe)
```

The most important part in topology-based pathway enrichment analysis is that we need pathways in the network representations.
The **KEGGgraph** can read network data of KEGG pathways and convert it to proper format.

```{r}
library(KEGGgraph)
df = parseKGML2DataFrame(system.file("extdata/hsa04010.xml", package = "KEGGgraph"))
head(df)
df[, 1] = gsub("^hsa:", "", df[, 1])
df[, 2] = gsub("^hsa:", "", df[, 2])
```

We only need the first two columns (from and to) to construct the network.

```{r}
library(igraph)
g = graph_from_data_frame(df[, 1:2])
g
```

The **igraph** package provides many different centrality methods. Here we take `degree()` as an example:

```{r}
d = degree(g)
```

Make sure all genes in `d` are within `universe`.

```{r}
d = d[intersect(names(d), universe)]
```

The statistic for the gene set is $\sum_i^n I(G_i) \cdot d_i$

```{r}
cn = intersect(genes, names(d))
s = sum(d[cn])
```

or

```r
s = sum( (names(d) %in% genes)*d )
```

In the random permutation, we randomly pick `length(genes)` as differential genes.


```{r}
s_random = numeric(1000)
for(i in 1:1000) {
    genes_random = sample(universe, length(genes))
    cn = intersect(genes_random, names(d))
    s_random[i] = sum(d[cn])
}
```

P-value is calculated in the normal way.

```{r}
hist(s_random)
abline(v = s, col = 2)
sum(s_random > s)/length(s_random)
```

## GSEA-extension

We only consider the gene permutation process. We generate a vector of gene scores.

```{r}
score = structure(rnorm(length(universe)), names = universe)
```

As `d` has already been restricted within `universe`, the set-level statistic is calculated as $\frac{1}{n}\sum_i^n |s_i|\cdot d_i$

```{r}
s = mean(abs(score[names(d)])*d)
```

In the gene permutation, the differential scores are randomly assigned to genes.

```{r}
s_random = numeric(1000)
for(i in 1:1000) {
    score_random = sample(score, length(d))
    s_random[i] = mean(abs(score_random)*d)
}
hist(s_random)
abline(v = s, col = 2)
sum(s_random > s)/length(s_random)
```


**Question**: can you do a two-sided test?

## Helper functions

Two helper funtions are provided:

```{r}
ora_kegg_topology(genes, universe, centrality = igraph::degree, verbose = FALSE) |> head()
gsea_kegg_topology(score, centrality = igraph::degree, verbose = FALSE) |> head()
```

## CePa

The **CePa** package has not been updated for ~10 years. 


It will try six different centralities:

- equal.weight
- in.degree
- out.degree
- node betweenness
- in.reach
- out.reach


`dif` and `bk` are two gene vectors, `pc` is in a special format.

```{r, eval = FALSE}
# around 5min
res = cepa.ora.all(dif = gene.list$dif, bk = gene.list$bk, pc = PID.db$NCI)
```

```{r, echo = FALSE}
res = readRDS(system.file("extdata", "cepa_res.rds", package = "GSEAtraining"))
```

Heatmap of adjusted p-values:

```{r}
plot(res, adj.method = "BH", only.sig = TRUE)
```


Advantages:

1. use multiple centrality metrics,
2. node-level instead of gene-level
3. also include non-gene nodes.

