---
title: "Topic 2-06: Compare ORA and GSEA"
author: "Zuguang Gu <z.gu@dkfz.de>"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Topic 2-06: Compare ORA and GSEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo = FALSE}
library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    warning = FALSE,
    fig.width = 8,
    fig.height = 8,
    fig.align = "center"
)
```


We apply ORA and GSEA on the same p53 datasets and compare the results.

```{r}
lt = readRDS(system.file("extdata", "p53_expr.rds", package = "GSEAtutorial"))
expr = lt$expr
condition = lt$condition

ln = strsplit(readLines(system.file("extdata", "c2.symbols.gmt", package = "GSEAtutorial")), "\t")
gs = lapply(ln, function(x) x[-(1:2)])
names(gs) = sapply(ln, function(x) x[1])
```

**ORA**

We need to perform t-test or similar DE test to extract DE genes:

We apply t-test to every gene.

```{r}
p = apply(expr, 1, function(x) {
    x1 = x[condition == "WT"]
    x2 = x[condition == "MUT"]
    
    t.test(x1, x2)$p.value
})
```


Faster row-wise t-test: 

```{r}
library(genefilter)
tdf = rowttests(expr, factor(condition))  # the second must be a "factor"
tdf$fdr = p.adjust(tdf$p.value, "BH")
sum(tdf$fdr < 0.05)  # number of diff genes
```

It seems there are not too many diff genes. We look at the distribution of the t-statistics:

```{r}
plot(sort(tdf$statistic))
```

Instead of setting a cutoff for FDR, we set a cutoff for t-statistics, just to get
enough number of diff genes.

```{r}
sum(abs(tdf$statistic) > 2)
```

```{r}
diff = rownames(expr)[abs(tdf$statistic) > 2]
```

First we perform ORA, using GO BP gene sets:

```{r, message = FALSE}
tb_ora = ora(diff, universe = rownames(expr), gene_sets = gs)
```


Then we perform GSEA, using signal-to-noise ratios as gene scores. 

```{r}
library(matrixStats)
m1 = expr[, condition == "WT"]  # only samples in group 1
m2 = expr[, condition == "MUT"]  # only samples in group 2
s = (rowMeans(m1) - rowMeans(m2))/(rowSds(m1) + rowSds(m2))
s = sort(s, decreasing = TRUE)

tb_gsea = fgsea_wrapper(s, gs)
```

We compare the two significant GO lists:

```{r}
library(eulerr)
plot(euler(list(ORA = tb_ora$gene_set[tb_ora$p_value < 0.05], 
                GSEA = tb_gsea$gene_set[tb_gsea$p_value < 0.05])
          ), quantities = TRUE)
```

```{r, fig.width = 10, fig.height = 5}
rownames(tb_ora) = tb_ora$gene_set
rownames(tb_gsea) = tb_gsea$gene_set
cn = intersect(rownames(tb_ora), rownames(tb_gsea))

par(mfrow = c(1, 2))
plot(tb_ora[cn, "p_value"], tb_gsea[cn, "p_value"])
plot(tb_ora[cn, "log2fe"], log2(abs(tb_gsea[cn, "nes"])))
```



```{r}
tb_ora1 = ora_go(convert_to_entrez_id(diff))
tb_ora2 = ora_go(convert_to_entrez_id(diff), universe = convert_to_entrez_id(rownames(expr)))
tb_gsea = fgsea_go(convert_to_entrez_id(s))

lt_GO = list(ORA1 = tb_ora1$gene_set[order(tb_ora1$p_value)[1:100]], 
              ORA2 = tb_ora2$gene_set[order(tb_ora2$p_value)[1:100]], 
              GSEA = tb_gsea$gene_set[order(tb_gsea$p_value)[1:100]])
plot(euler(lt_GO), quantities = TRUE)
```


```{r, fig.width = 10, fig.height = 5}
library(simplifyEnrichment)
simplifyGOFromMultipleLists(lt_GO)
```
